---
title: "Mean Annual Discharge per Segment"
output: bookdown::html_vignette2
vignette: >
  %\VignetteIndexEntry{Mean Annual Discharge per Segment}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.width = 7,
  out.width = "100%",
  dpi = 150
)
```

```{r load}
library(wet)
library(ggplot2)
sf::sf_use_s2(FALSE)

vd <- function(f) system.file("vignette-data", f, package = "wet")
v <- readRDS(vd("segment_values.rds"))
prov <- v$provenance
map <- readRDS(vd("segment_map.rds"))
bc <- readRDS(vd("station_map.rds"))$bc

# colours come from the gq registries, so no hex literal appears in this
# vignette; assert each class set, because a malformed CSV row adds a layer silently
reg <- gq::gq_reg_custom(system.file("cartography", "wet_segment.csv", package = "wet"))
reg_main <- gq::gq_reg_main()
classes <- function(layer, expect) {
  x <- gq::gq_tmap_classes(reg$layers[[layer]])
  stopifnot(identical(names(x$values), expect))
  list(values = x$values, labels = stats::setNames(x$labels, names(x$values)))
}
pal_mad <- classes("mad_class", paste0("m", 1:6))
pal_map <- classes("segment_map", c("group", "outline", "key"))
pal_skill <- classes("skill_nesting", c("headwater", "nested", "mae"))
pal_prod <- classes("product", c("obs", "wb", "fwapg"))
pal_par <- classes("parity", c("segment", "one_to_one"))
pal_err <- classes("gauge_error", paste0("e", 1:4))
lake_sty <- reg_main$layers$lake
town_sty <- reg_main$layers$town

# discharge classes: labels that restate the breaks are built from them
mad_breaks <- c(0, 0.1, 0.5, 2, 10, 50, Inf)
mad_width <- c(0.15, 0.25, 0.4, 0.6, 0.85, 1.1)
fmt_q <- function(x) format(x, drop0trailing = TRUE, trim = TRUE)
pal_mad$labels[] <- c(paste("Under", fmt_q(mad_breaks[2])),
                      paste0(fmt_q(mad_breaks[2:5]), "–", fmt_q(mad_breaks[3:6])),
                      paste(fmt_q(mad_breaks[6]), "or more"))
mad_class <- function(q) factor(names(pal_mad$values)[findInterval(q, mad_breaks)], names(pal_mad$values))
err_breaks <- c(-Inf, -20, 0, 20, Inf)
pal_err$labels[] <- c(paste0("More than ", -err_breaks[2], "% under"), paste0("0–", -err_breaks[2], "% under"),
                      paste0("0–", err_breaks[4], "% over"), paste0("More than ", err_breaks[4], "% over"))
err_class <- function(e) factor(names(pal_err$values)[findInterval(e, err_breaks)], names(pal_err$values))

group_name <- stats::setNames(map$groups$watershed_group_name, map$groups$watershed_group_code)
group_km2 <- stats::setNames(map$groups$area_km2, map$groups$watershed_group_code)
theme_set(theme_minimal(base_size = 10) +
            theme(legend.position = "bottom", panel.grid.minor = element_blank()))
pct <- function(x) paste0(round(x), "%")
signed <- function(x) sub("^-", "−", sprintf("%+.0f%%", x))
num <- function(x, d = 0) format(round(x, d), big.mark = ",", nsmall = d)
and_text <- function(x) sub(", ([^,]*)$", " and \\1", paste(x, collapse = ", "))
to_title <- function(x) {
  vapply(strsplit(tolower(x), " "), function(w) {
    big <- !w %in% c("at", "the", "near", "of", "in", "above", "below", "north")
    w[big] <- paste0(toupper(substring(w[big], 1, 1)), substring(w[big], 2))
    paste(w, collapse = " ")
  }, "")
}
```

wet estimates mean annual discharge (MAD) for every stream segment of BC's
Freshwater Atlas, in two ways. The first follows fwapg: PCIC's VIC-GL runoff
and baseflow, averaged over 1981–2010 and accumulated upstream. It covers only
the Peace, Fraser and Columbia. The second is wet's own open water balance,
precipitation less evapotranspiration, adjusted by zone against HYDAT gauges.
It covers the province. This vignette shows both on two watershed groups: the
`r group_name[["SALR"]]` (`r num(group_km2[["SALR"]])` km²), a headwater
Fraser group, and the `r group_name[["BULK"]]` (`r num(group_km2[["BULK"]])`
km²), in the Skeena, where fwapg has no discharge.

## Reproducing fwapg

```{r parity-data}
p <- v$parity
n_par <- nrow(p)
same5 <- mean(round(p$mad_m3s_wet, 5) == round(p$mad_m3s_fwapg, 5))
stopifnot(n_par == prov$fwapg_discharge_rows[["SALR"]], same5 == 1)   # the prose says "every"
# fwapg has no discharge in the Bulkley: the intro and the limits say so
stopifnot(prov$fwapg_discharge_rows[["BULK"]] == 0)
cap_par <- sprintf(paste(
  "Mean annual discharge on the %s %s segments fwapg gives a value: its stored value against wet's rebuild",
  "of it from PCIC VIC-GL, 1981–2010, with the 1:1 line dashed. Both axes are logarithmic."
), num(n_par), group_name[["SALR"]])
```

fwapg samples each watershed polygon at the PCIC cell under its centroid. It
then averages over the polygon and everything upstream of it, weighted by
area. wet rebuilds that number with its own code. On every one of the
`r num(n_par)` `r group_name[["SALR"]]` segments that fwapg gives a value,
it matches fwapg to the five decimals fwapg stores.

```{r parity, fig.width = 4.5, fig.height = 4.5, out.width = "65%", fig.cap = cap_par}
q_lab <- function(x) format(x, scientific = FALSE, drop0trailing = TRUE, trim = TRUE)
ggplot(p, aes(mad_m3s_fwapg, mad_m3s_wet)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = pal_par$values[["one_to_one"]],
              linewidth = 0.4) +
  geom_point(colour = pal_par$values[["segment"]], size = 0.6, alpha = 0.4) +
  scale_x_log10(labels = q_lab) +
  scale_y_log10(labels = q_lab) +
  coord_equal() +
  labs(x = "fwapg (m³/s)", y = "wet (m³/s)")
```

## Sampling small watersheds

```{r sampling-data}
sm <- v$sampling
sm$change <- 100 * (sm$mm_area / sm$mm_centroid - 1)
area_cut <- c(0, 10, 100, Inf)
sm$size <- cut(sm$upstream_area_km2, area_cut)
moved <- tapply(abs(sm$change) > 5, sm$size, mean)
q99 <- tapply(abs(sm$change), sm$size, stats::quantile, 0.99)
# the prose: small watersheds move, large ones barely
stopifnot(moved[[1]] > moved[[3]], q99[[1]] > 5 * q99[[3]])
# PCIC's 0.0625-degree cell, in km at the group's latitude
lat <- sf::st_coordinates(sf::st_transform(sf::st_centroid(sf::st_geometry(
  map$groups[map$groups$watershed_group_code == "SALR", ])), 4326))[, 2]
cell_ns <- 0.0625 * 111.2
cell_ew <- cell_ns * cos(lat * pi / 180)
cap_samp <- sprintf(paste(
  "Change in mean annual runoff from area-weighted sampling, against centroid sampling, for",
  "the %s %s watersheds that carry a stream. Upstream area is on a logarithmic axis."
), num(nrow(sm)), group_name[["SALR"]])
```

A PCIC cell is 0.0625°, about `r round(cell_ns)` by `r round(cell_ew)` km
here, so the centroid rule hands a small watershed the value of one cell. wet can instead weight every cell by the
area it covers. Under `r area_cut[2]` km², `r pct(100 * moved[[1]])` of
watersheds move by more than 5%, by up to `r pct(q99[[1]])` at the 99th
percentile. Above `r area_cut[3]` km², the 99th percentile is
`r format(round(q99[[3]], 1), nsmall = 1)`%, because a large basin averages
many cells either way.

```{r sampling, fig.height = 4, fig.cap = cap_samp}
ggplot(sm, aes(upstream_area_km2, change)) +
  geom_hline(yintercept = 0, colour = pal_par$values[["one_to_one"]], linewidth = 0.3) +
  geom_point(colour = pal_par$values[["segment"]], size = 0.6, alpha = 0.4) +
  scale_x_log10(labels = function(x) format(x, big.mark = ",", scientific = FALSE, drop0trailing = TRUE)) +
  labs(x = "Upstream area (km²)", y = "Change (%)")
```

## Two groups, two products

```{r map-data}
seg <- merge(map$segments, v$segments, by = c("linear_feature_id", "watershed_group_code"))
stopifnot(nrow(seg) == nrow(map$segments))
# the Salmon River is drawn from PCIC through wet (fwapg's numbers), the
# Bulkley from the open water balance, the only product there
seg$mad <- ifelse(seg$watershed_group_code == "SALR", seg$mad_pcic_m3s, seg$mad_wb_m3s)
seg$class <- mad_class(seg$mad)
n_no_value <- sum(is.na(seg$mad))
stopifnot(identical(is.na(seg$mad), is.na(seg$mad_wb_m3s)))   # the bullet says neither product has them
order_no_value <- range(seg$stream_order[is.na(seg$mad)])
n_total <- table(seg$watershed_group_code)
seg <- seg[!is.na(seg$mad), ]
seg <- seg[order(seg$class), ]   # large rivers drawn last, on top

sk <- v$skill
gauges <- sf::st_transform(sf::st_as_sf(sk[sk$watershed_group_code == "BULK", ], coords = c("lon", "lat"),
                                        crs = 4326), 3005)
gauges$class <- err_class(gauges$err_pct)
stopifnot(nrow(gauges) > 0)

panel_w <- 4.6
panel_h <- 6.4
frame <- function(code) {
  g <- map$groups[map$groups$watershed_group_code == code, ]
  bb <- sf::st_bbox(sf::st_buffer(g, 4000))
  bb <- gq::gq_bbox_aspect(bb, asp = panel_w / panel_h)
  list(group = g, bb = bb, lim_x = bb[c("xmin", "xmax")], lim_y = bb[c("ymin", "ymax")],
       dx = bb[["xmax"]] - bb[["xmin"]], dy = bb[["ymax"]] - bb[["ymin"]])
}

draw_map <- function(code, gauges = NULL) {
  f <- frame(code)
  s <- seg[seg$watershed_group_code == code, ]
  lakes <- map$lakes[map$lakes$watershed_group_code == code, ]
  sb_km <- max(gq::gq_scale_breaks(f$bb))
  sb <- data.frame(x = f$lim_x[[2]] - 0.05 * f$dx - sb_km * 1000, y = f$lim_y[[1]] + 0.04 * f$dy)
  sb$xend <- sb$x + sb_km * 1000
  key <- ggplot() +
    geom_sf(data = bc, fill = "white", colour = pal_map$values[["outline"]], linewidth = 0.2) +
    geom_sf(data = f$group, fill = pal_map$values[["key"]], colour = NA) +
    geom_sf(data = sf::st_centroid(sf::st_geometry(f$group)), shape = 1, size = 6, stroke = 0.6,
            colour = pal_map$values[["key"]]) +
    theme_void() +
    theme(panel.background = element_rect(fill = "white", colour = pal_map$values[["outline"]],
                                          linewidth = 0.3))
  key_w <- 0.34 * f$dx
  key_h <- key_w * 1.05
  out <- ggplot() +
    geom_sf(data = f$group, fill = pal_map$values[["group"]], colour = pal_map$values[["outline"]],
            linewidth = 0.4) +
    geom_sf(data = lakes, fill = lake_sty$fill$color, colour = lake_sty$stroke$color, linewidth = 0.15) +
    geom_sf(data = s, aes(colour = class, linewidth = class)) +
    scale_colour_manual(values = pal_mad$values, labels = pal_mad$labels,
                        name = "Mean annual\ndischarge (m³/s)") +
    scale_linewidth_manual(values = stats::setNames(mad_width, names(pal_mad$values)),
                           labels = pal_mad$labels, name = "Mean annual\ndischarge (m³/s)") +
    geom_segment(data = sb, aes(x = x, xend = xend, y = y, yend = y), linewidth = 1) +
    geom_text(data = sb, aes(x = (x + xend) / 2, y = y, label = paste(sb_km, "km")), vjust = -0.7, size = 3.2) +
    annotation_custom(ggplotGrob(key), xmin = f$lim_x[[2]] - key_w - 0.02 * f$dx,
                      xmax = f$lim_x[[2]] - 0.02 * f$dx,
                      ymin = f$lim_y[[2]] - key_h - 0.02 * f$dy, ymax = f$lim_y[[2]] - 0.02 * f$dy)
  pl <- map$places[map$places$watershed_group_code == code, ]
  if (nrow(pl)) {
    # a name sits left of its town, or right of it where the left would leave
    # the frame; a few pixels above, so a gauge at the town stays visible
    xy <- sf::st_coordinates(pl)
    w <- nchar(pl$name) * 3 * 0.6 / (25.4 * panel_w) * f$dx   # label width at size 3, in metres
    left <- xy[, 1] - 0.012 * f$dx - w > f$lim_x[[1]] + 0.01 * f$dx
    lab <- data.frame(x = xy[, 1] + ifelse(left, -0.012, 0.012) * f$dx, y = xy[, 2] + 0.012 * f$dy,
                      label = pl$name, hjust = ifelse(left, 1, 0))
    out <- out +
      geom_point(data = data.frame(x = xy[, 1], y = xy[, 2]), aes(x, y), size = 1.2,
                 colour = town_sty$label$color) +
      geom_label(data = lab, aes(x, y, label = label, hjust = hjust), colour = town_sty$label$color,
                 fontface = "bold", size = 3, linewidth = 0, fill = "white", alpha = 0.8,
                 label.padding = unit(1.2, "pt"))
  }
  if (!is.null(gauges)) {
    out <- out +
      geom_sf(data = gauges, aes(fill = class), shape = 21, colour = "black", size = 2.8, stroke = 0.4) +
      scale_fill_manual(values = pal_err$values, labels = pal_err$labels, name = "Gauge, held-out\nerror")
  }
  out +
    coord_sf(xlim = f$lim_x, ylim = f$lim_y, expand = FALSE, datum = NA) +
    theme_void() +
    theme(legend.position = "right", legend.key.width = unit(14, "pt"),
          panel.border = element_rect(fill = NA, colour = pal_map$values[["outline"]], linewidth = 0.3))
}

# SALR: both products, and the gauges that arbitrate between them
ss <- v$segments[v$segments$watershed_group_code == "SALR" & !is.na(v$segments$mad_wb_m3s), ]
ratio_salr <- stats::median(ss$mad_wb_m3s / ss$mad_pcic_m3s)
ga <- v$gauges_salr
ga$wb_err <- 100 * (ga$wb_mm / ga$obs_mm - 1)
ga$fw_err <- 100 * (ga$fwapg_mm / ga$obs_mm - 1)
outlet <- ga[ga$holds_salr, ]
# the balance's share of the gap at each gauge, in log terms like the ratio:
# log(wb / fwapg) = log(wb / obs) + log(obs / fwapg)
ga$wb_share <- log(ga$wb_mm / ga$obs_mm) / log(ga$wb_mm / ga$fwapg_mm)
close <- abs(ga$fw_err) <= 5
# the prose: the balance reads high at every gauge; at the small basins fwapg is
# within 5% and the gap is nearly all the balance's; at the large ones, the
# outlet among them, fwapg is low and the two share it
stopifnot(ratio_salr > 1, all(ga$wb_err > 0), nrow(outlet) == 1,
          any(close), any(!close), max(ga$area_km2[close]) < min(ga$area_km2[!close]),
          all(ga$wb_share[close] > 0.9), all(ga$fw_err[!close] < -5),
          all(ga$wb_share[!close] > 0.25 & ga$wb_share[!close] < 0.75), !close[ga$holds_salr])
gauge_tab <- data.frame(
  Gauge = sprintf("%s %s", ga$station_number, to_title(ga$station_name)),
  `Area (km²)` = num(ga$area_km2),
  `Observed (mm)` = num(ga$obs_mm),
  `Water balance (held out)` = signed(ga$wb_err),
  fwapg = signed(ga$fw_err),
  check.names = FALSE
)
gauge_tab <- gauge_tab[order(!ga$holds_salr, ga$area_km2), ]

nb <- nrow(gauges)
bulk_zones <- sort(unique(sk$zone[sk$watershed_group_code == "BULK"]))
bulk_mae <- mean(abs(gauges$err_pct))
cap_salr <- sprintf(paste(
  "Mean annual discharge on %s of the %s %s segments of order %d or more, from PCIC VIC-GL",
  "through wet (fwapg's values). The keymap marks the group in BC."
), num(sum(seg$watershed_group_code == "SALR")), num(n_total[["SALR"]]), group_name[["SALR"]],
prov$min_order)
cap_bulk <- sprintf(paste(
  "Mean annual discharge on %s of the %s %s segments of order %d or more, from the open water",
  "balance, and the %d calibration gauges in the group, filled by the held-out error of the",
  "balance at each."
), num(sum(seg$watershed_group_code == "BULK")), num(n_total[["BULK"]]), group_name[["BULK"]],
prov$min_order, nb)
```

Both products cover the `r group_name[["SALR"]]`, and they disagree there.
The open water balance gives a median `r format(round(ratio_salr, 2), nsmall = 2)`
times the PCIC value per segment. The gauges nearby show where each misses.
The balance runs high at all `r nrow(ga)` of them, by
`r signed(min(ga$wb_err))` to `r signed(max(ga$wb_err))`. At the
`r sum(close)` smallest basins fwapg is within 5%, so the gap there is the
balance's. At the `r sum(!close)` larger ones fwapg runs
`r signed(max(ga$fw_err[!close]))` to `r signed(min(ga$fw_err[!close]))`, and
the two share it. At `r outlet$station_number`, which drains the group, they
bracket the observation: `r signed(outlet$wb_err)` against
`r signed(outlet$fw_err)`. The map below shows PCIC's values. The balance's
errors in the table are out of sample (blocked cross-validation).

```{r gauges-salr}
knitr::kable(gauge_tab, align = "lrrrr", row.names = FALSE,
             caption = sprintf(paste(
               "Calibration gauges within %d km of the %s group, and its outlet gauge (first row):",
               "mean annual runoff observed in 1981–2010, and each product's error against it."
             ), prov$near_km, group_name[["SALR"]]))
```

```{r map-salr, fig.width = panel_w + 1.8, fig.height = panel_h, fig.cap = cap_salr}
draw_map("SALR")
```

fwapg has `r prov$fwapg_discharge_rows[["BULK"]]` discharge rows in the
`r group_name[["BULK"]]`, which lies outside PCIC's coverage. The open water
balance supplies a value for every segment that has a watershed. The
`r nb` gauges in the group are part of the balance's validation. Their
held-out error averages `r pct(bulk_mae)` in absolute terms.

```{r map-bulk, fig.width = panel_w + 1.8, fig.height = panel_h, fig.cap = cap_bulk}
draw_map("BULK", gauges)
```

## Skill out of sample

```{r skill-data}
sk$ratio <- sk$mod / sk$obs
mae_of <- function(i) mean(abs(sk$err_pct[i]))
mae_all <- mae_of(TRUE)
mae_head <- mae_of(sk$nesting == "headwater")
mae_nest <- mae_of(sk$nesting == "nested")
in20 <- mean(abs(sk$err_pct) <= 20)
zone_mae <- sort(tapply(abs(sk$err_pct), sk$zone, mean))
sk$zone <- factor(sk$zone, names(zone_mae))
zl <- data.frame(zone = factor(names(zone_mae), names(zone_mae)), mae = zone_mae,
                 n = as.vector(table(sk$zone)[names(zone_mae)]))
worst <- names(zone_mae)[length(zone_mae)]
# the prose: nested gauges score better than headwater, and the worst zone is a
# semi-arid interior plateau (15, 17, 23 or 24: research/water_balance_method.md)
# that stands out
stopifnot(mae_nest < mae_head, sum(zone_mae > 2 * mae_all) == 1, worst %in% c("15", "17", "23", "24"))
ratio_breaks <- c(0.5, 0.75, 1, 1.5, 2, 3, 5)
cap_skill <- sprintf(paste(
  "Modelled against observed mean annual runoff at the %d calibration gauges, each predicted",
  "with its own sub-sub-drainage held out of the fit, by hydrologic zone. Zones are ordered by",
  "mean absolute error, printed at the right with the number of gauges."
), nrow(sk))
```

The balance's coefficients are fitted on HYDAT gauges, so its skill is
scored on gauges the fit did not see. Each WSC sub-sub-drainage is held out
in turn and predicted from the rest. Over `r nrow(sk)` gauges, the mean
absolute error in annual runoff is `r pct(mae_all)`, and `r pct(100 * in20)`
fall within ±20%. Headwater gauges score `r pct(mae_head)`, and nested ones,
whose basins hold other gauges, score `r pct(mae_nest)`. Zone `r worst`, a
semi-arid interior plateau, is the outlier at `r pct(zone_mae[[worst]])`.
The `r group_name[["BULK"]]` gauges fall in zones `r and_text(bulk_zones)`.

```{r skill, fig.height = 6.5, fig.cap = cap_skill}
x_lab <- max(ratio_breaks) * 1.6
ggplot(sk, aes(ratio, zone)) +
  geom_vline(xintercept = 1, colour = pal_par$values[["one_to_one"]], linewidth = 0.3) +
  geom_point(aes(colour = nesting), size = 1.6, alpha = 0.8) +
  geom_text(data = zl, aes(x = x_lab, y = zone, label = sprintf("%s%% (%d)", round(mae), n)),
            hjust = 1, size = 2.8, colour = pal_skill$values[["mae"]]) +
  scale_x_log10(breaks = ratio_breaks, labels = sub("^\\+0%$", "0%", signed(100 * (ratio_breaks - 1))),
                limits = c(min(sk$ratio, ratio_breaks), x_lab)) +
  scale_colour_manual(values = pal_skill$values[c("headwater", "nested")],
                      labels = pal_skill$labels[c("headwater", "nested")], name = NULL) +
  labs(x = "Held-out error in mean annual runoff (logarithmic)", y = "Hydrologic zone")
```

## The recipe

This is the PCIC path for one headwater group, as `scripts/mad_parity.R` runs
it. It needs fwapg and the PCIC server. The open water balance is a pipeline of
scripts (`scripts/wb_inputs.R` to `scripts/wb_output.R`; see the package's
`CLAUDE.md`). Both run on cached outputs here (see Cached inputs, below).

```{r recipe, echo = TRUE, eval = FALSE}
library(wet)
conn <- DBI::dbConnect(RPostgres::Postgres(), dbname = "fwapg")

# PCIC VIC-GL runoff and baseflow over the group's extent, 1981-2010
bbox <- c(-124.12, 54.22, -123.12, 55.11)
runoff <- terra::rast(wet_pcic_fetch("RUNOFF", bbox, "1981-01-01", "2010-12-31"))
baseflow <- terra::rast(wet_pcic_fetch("BASEFLOW", bbox, "1981-01-01", "2010-12-31"))
mad_cell <- wet_runoff_annual(runoff) + wet_runoff_annual(baseflow)   # mm/yr per cell

# every watershed polygon in the Fraser; those inside the extent get a cell value
ws <- wet_ws_fetch(conn, "100")
in_bb <- ws$lon >= bbox[1] & ws$lon <= bbox[3] & ws$lat >= bbox[2] & ws$lat <= bbox[4]
cells <- wet_ws_sample(mad_cell, ws[in_bb, c("watershed_feature_id", "lon", "lat")], "centroid")
up <- wet_upstream_mean(ws, cells, "total", irregular_pairs = wet_upstream_irregular(conn, "100"))
up$mad_m3s <- wet_mm_to_m3s(up$value, up$upstream_area_m2)
```

## What it does not show

```{r limits}
stale <- prov$upstream_area_100
# the bullet: stale polygons exist in the Fraser, and none moves a SALR segment
stopifnot(stale[["stale"]] > 0, prov$salr_stale_segments == 0)
```

- **fwapg's stored upstream area is a stale snapshot.** In the Fraser,
  `r num(stale[["stale"]])` of `r num(stale[["polygons"]])` polygons carry a
  stored area that differs from the live accumulation. None of them changes
  a `r group_name[["SALR"]]` segment, so the parity above does not test it. wet
  uses the accumulated area except to reproduce fwapg.
- **PCIC covers only the Peace, Fraser and Columbia.** Elsewhere, as in the
  `r group_name[["BULK"]]`, the open water balance is the only estimate. A
  model that classifies habitat on fwapg's discharge gets nothing there.
- **The zone adjustment passes its gate only under a variant added later.**
  The pre-set specification gave sparsely gauged zones one shared level and
  failed. The "no adjustment" variant that passes was offered after that
  result, so the skill above is mildly optimistic
  (`research/water_balance_method.md`).
- **Segments without a watershed.** `r n_no_value` segments of order
  `r order_no_value[1]` to `r order_no_value[2]` have no watershed polygon in the
  Freshwater Atlas lookup, so neither product gives them a value. They are
  left off the maps, which leaves short gaps in some streams.

**Cached inputs.** PCIC VIC-GL through `scripts/mad_parity.R`; the open
water balance from province run `r prov$province_run` (annual AET
`r prov$aet`, HYDAT release `r substr(prov$hydat_release, 1, 10)`); built
`r paste(as.integer(format(prov$built, "%d")), format(prov$built, "%B %Y"))` at commit `r prov$wet_commit`. Regenerated by
`data-raw/segment_vignette_data.R`. Map layers from the Freshwater Atlas
(fwapg) and BC Geographic Names, by `data-raw/segment_vignette_map.R`.

```{r wordcount, include = FALSE}
f <- knitr::current_input()
if (!is.null(f)) {
  x <- readLines(f, warn = FALSE)
  y <- utils::tail(x, -utils::head(which(x == "---"), 2)[2])   # drop the YAML header
  fence <- cumsum(grepl("^```", y)) %% 2 == 1
  y <- y[!fence & !grepl("^```", y)]
  y <- gsub("`r [^`]*`", "", y)          # inline code is not body prose
  y <- y[!grepl("^\\s*$", y)]
  n <- sum(lengths(strsplit(trimws(y), "\\s+")))
  message("body prose: ", n, " words")
  cap <- 800
  if (n > cap) stop("body prose is ", n, " words; the cap is ", cap)
}
```
