---
title: "Mean Annual Discharge per Segment"
output: bookdown::html_vignette2
vignette: >
  %\VignetteIndexEntry{Mean Annual Discharge per Segment}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.width = 7,
  out.width = "100%",
  dpi = 150
)
```

```{r load}
library(wet)
library(ggplot2)
sf::sf_use_s2(FALSE)

vd <- function(f) system.file("vignette-data", f, package = "wet")
v <- readRDS(vd("segment_values.rds"))
prov <- v$provenance
map <- readRDS(vd("segment_map.rds"))
bc <- readRDS(vd("station_map.rds"))$bc

# colours come from the gq registries, so no hex literal appears in this
# vignette; assert each class set, because a malformed CSV row adds a layer silently
reg <- gq::gq_reg_custom(system.file("cartography", "wet_segment.csv", package = "wet"))
reg_main <- gq::gq_reg_main()
classes <- function(layer, expect) {
  x <- gq::gq_tmap_classes(reg$layers[[layer]])
  stopifnot(identical(names(x$values), expect))
  list(values = x$values, labels = stats::setNames(x$labels, names(x$values)))
}
pal_mad <- classes("mad_class", paste0("m", 1:6))
pal_map <- classes("segment_map", c("group", "outline", "key"))
pal_skill <- classes("skill_nesting", c("headwater", "nested", "mae"))
pal_par <- classes("parity", c("segment", "one_to_one"))
pal_err <- classes("gauge_error", paste0("e", 1:4))
pal_prov <- classes("province", c("coverage", "zone", "group"))
pal_ctx <- classes("context", "river")
lake_sty <- reg_main$layers$lake
town_sty <- reg_main$layers$town

# discharge classes: labels that restate the breaks are built from them
mad_breaks <- c(0, 0.1, 0.5, 2, 10, 50, Inf)
mad_width <- c(0.15, 0.25, 0.4, 0.6, 0.85, 1.1)
fmt_q <- function(x) format(x, drop0trailing = TRUE, trim = TRUE)
pal_mad$labels[] <- c(paste("Under", fmt_q(mad_breaks[2])),
                      paste0(fmt_q(mad_breaks[2:5]), "–", fmt_q(mad_breaks[3:6])),
                      paste(fmt_q(mad_breaks[6]), "or more"))
mad_class <- function(q) factor(names(pal_mad$values)[findInterval(q, mad_breaks)], names(pal_mad$values))
# the one error scale: every figure, map and table below uses these classes
err_cut <- 20
pal_err$labels[] <- c(paste0("More than ", err_cut, "% under"), paste0("0–", err_cut, "% under"),
                      paste0("0–", err_cut, "% over"), paste0("More than ", err_cut, "% over"))
# within ±err_cut (the table's test, ends included) is the two middle classes
err_class <- function(e) {
  k <- ifelse(e < -err_cut, 1L, ifelse(e < 0, 2L, ifelse(e <= err_cut, 3L, 4L)))
  factor(names(pal_err$values)[k], names(pal_err$values))
}
err_scale <- function(name = "Error against\nthe gauge") {
  # drop = FALSE keeps a class no gauge falls in; override.aes gives it a symbol
  scale_fill_manual(values = pal_err$values, labels = pal_err$labels, name = name, drop = FALSE,
                    guide = guide_legend(override.aes = list(shape = 21, size = 2.6, colour = "black")))
}

group_name <- stats::setNames(map$groups$watershed_group_name, map$groups$watershed_group_code)
group_km2 <- stats::setNames(map$groups$area_km2, map$groups$watershed_group_code)
theme_set(theme_minimal(base_size = 10) +
            theme(legend.position = "bottom", panel.grid.minor = element_blank()))
pct <- function(x) paste0(round(x), "%")
pct1 <- function(x) paste0(format(round(x, 1), nsmall = 1), "%")
signed <- function(x) sub("^-", "−", sprintf("%+.0f%%", x))
num <- function(x, d = 0) format(round(x, d), big.mark = ",", nsmall = d)
and_text <- function(x) sub(", ([^,]*)$", " and \\1", paste(x, collapse = ", "))
to_title <- function(x) {
  vapply(strsplit(tolower(x), " "), function(w) {
    big <- !w %in% c("at", "the", "near", "of", "in", "above", "below", "north")
    w[big] <- paste0(toupper(substring(w[big], 1, 1)), substring(w[big], 2))
    paste(w, collapse = " ")
  }, "")
}

# every station the text or a table names must sit inside a map's frame; each
# is recorded where it is named, and each map records what its frame holds
named <- character()
drawn <- character()
in_frame <- function(pts, bb) {
  xy <- sf::st_coordinates(pts)
  pts$station_number[xy[, 1] > bb[["xmin"]] & xy[, 1] < bb[["xmax"]] &
                       xy[, 2] > bb[["ymin"]] & xy[, 2] < bb[["ymax"]]]
}
as_pts <- function(x) sf::st_transform(sf::st_as_sf(x, coords = c("lon", "lat"), crs = 4326, remove = FALSE), 3005)

sk <- v$skill
# the major basins PCIC covers, by Water Survey drainage: 07E-07F the Peace,
# 08J-08M the Fraser, 08N the Columbia
basin_of <- function(st) {
  ifelse(substr(st, 1, 3) %in% c("07E", "07F"), "Peace",
         ifelse(substr(st, 1, 3) %in% paste0("08", c("J", "K", "L", "M")), "Fraser",
                ifelse(substr(st, 1, 3) == "08N", "Columbia", NA)))
}
sk$basin <- basin_of(sk$station_number)
```

wet gives every stream segment of British Columbia's Freshwater Atlas a mean
annual discharge: the average flow, in cubic metres per second, over
1981–2010. This article explains what that number is, the two estimates
behind it and what each suits, how close they come to measured flow, and
where they are weakest.

## What mean annual discharge is

```{r units}
ex_km2 <- 100
ex_mm <- 300
ex_q <- wet_mm_to_m3s(ex_mm, ex_km2 * 1e6)
stopifnot(abs(ex_q - 0.95) < 0.01)   # the prose's "about 0.95"
```

A stream's mean annual discharge is the volume of water that passes a point
in an average year, spread over the seconds in that year. It comes from
**runoff**: the depth of water, in millimetres a year, that the land upstream
yields to the stream once rain and snow have met evaporation and plant use.
Spread that depth over the area upstream and it becomes a volume. A
`r ex_km2` km² basin yielding `r ex_mm` mm a year passes about
`r format(round(ex_q, 2), nsmall = 2)` m³/s: 0.3 m over 100 million m² is 30
million m³ a year, and a year is 31.5 million seconds. Runoff in mm compares
basins of any size; discharge in m³/s is what a stream carries.
`wet_mm_to_m3s()` converts one to the other.

The terms used below:

- **[Freshwater Atlas](https://www2.gov.bc.ca/gov/content/data/geographic-data-services/topographic-data/freshwater)**
  (FWA): BC's mapped stream network. A *segment* is one piece of a stream
  line. Each segment drains a small *watershed* polygon, and the province is
  split into 246 *[watershed groups](https://catalogue.data.gov.bc.ca/dataset/freshwater-atlas-watershed-groups)*,
  such as the Salmon River and the Bulkley River below.
- **[Stream order](https://en.wikipedia.org/wiki/Strahler_number)**: 1 for a
  headwater stream, rising by one where two streams of the same order meet.
  The maps here draw order 3 and up.
- **[HYDAT](https://www.canada.ca/en/environment-climate-change/services/water-overview/quantity/monitoring/survey/data-products-services/national-archive-hydat.html)**:
  the Water Survey of Canada's archive of flow measured at gauges. A
  *calibration gauge* is one of the `r nrow(sk)` HYDAT gauges the water balance
  is fitted and scored on.
- **[PCIC hydrologic model output](https://services.pacificclimate.org/portal/hydro_model_out/map/)**:
  runoff and baseflow from the Pacific Climate Impacts Consortium's VIC-GL
  model, on a grid of 0.0625° cells. It covers the Peace, Fraser and Columbia.
- **[fwapg](https://github.com/smnorris/fwapg)**: an open database toolkit
  for the Freshwater Atlas. It publishes a mean annual discharge per segment
  from PCIC's output, and "fwapg" below means that estimate.
- **Open water balance**: wet's own estimate, after
  [Chapman et al. (2018)](https://www2.gov.bc.ca/assets/gov/environment/air-land-water/water/northeast-water-strategy/chapman_et_al-2018-jawra.pdf),
  the method behind the province's
  [water tools](https://www2.gov.bc.ca/gov/content/environment/air-land-water/water/water-science-data/water-data-tools).
  Precipitation less evapotranspiration per grid cell is averaged upstream,
  then adjusted by hydrologic zone against HYDAT gauges. Its code ships
  with wet, and its inputs are public.
- **[Hydrologic zone](https://catalogue.data.gov.bc.ca/dataset/bc-hydrologic-zones)**:
  one of BC's `r nrow(map$zones)` regions of similar runoff. The balance fits
  one adjustment per zone.
- **Held out**: scored at a gauge the fit did not see. wet holds out each
  Water Survey sub-sub-drainage in turn (the first four characters of a
  station number, such as 08EE), fits on the rest, and predicts the gauges it
  held out. This is *blocked cross-validation*: whole blocks are held out, so
  a gauge's neighbours on the same river cannot vouch for it.
- **Headwater and nested gauges**: a nested gauge's basin holds another
  gauge's; a headwater gauge's does not.
- **Error**: modelled minus observed runoff, as a percentage of observed. One
  scale runs through every figure: more than `r err_cut`% under,
  0–`r err_cut`% under, 0–`r err_cut`% over, more than `r err_cut`% over.

## Two estimates

```{r compare-data}
both <- sk[sk$in_pcic & !is.na(sk$fwapg_mm), ]
both$fw_err <- 100 * (both$fwapg_mm / both$obs - 1)
n_both <- nrow(both)
stopifnot(all(!is.na(both$basin)))   # every scored gauge is in the Peace, Fraser or Columbia
mae_wb <- mean(abs(both$err_pct))
mae_fw <- mean(abs(both$fw_err))
in20_wb <- mean(abs(both$err_pct) <= err_cut)
in20_fw <- mean(abs(both$fw_err) <= err_cut)
# the guidance: PCIC fits the gauges both cover more closely
stopifnot(mae_fw < mae_wb, in20_fw > in20_wb)
# the large-river row: gauges in PCIC's area on rivers fwapg leaves out, where
# the balance scores better than its average
dropped <- sk[sk$in_pcic & is.na(sk$fwapg_mm), ]
stopifnot(stats::median(dropped$area_km2) > stats::median(both$area_km2),
          mean(abs(dropped$err_pct)) < mae_wb)
o8 <- prov$fwapg_order8
stopifnot(o8[["valued"]] < 0.01 * o8[["segments"]])   # "rivers of order 8 and up are left out"
# grid cells in km at the Salmon River's latitude
lat_salr <- sf::st_coordinates(sf::st_transform(sf::st_centroid(sf::st_geometry(
  map$groups[map$groups$watershed_group_code == "SALR", ])), 4326))[[1, 2]]   # unnamed: c(ew = x) keeps x's name
cell_km <- function(deg) c(ns = deg * 111.2, ew = deg * 111.2 * cos(lat_salr * pi / 180))
km_pcic <- cell_km(0.0625)
km_wb <- cell_km(30 / 3600)
fmt_cell <- function(k, d) sprintf("about %s × %s km", format(round(k[["ns"]], d), nsmall = d),
                                   format(round(k[["ew"]], d), nsmall = d))
models <- data.frame(
  ` ` = c("What it is", "Where it has values", sprintf("Fit at the %d gauges both cover", n_both),
          "Largest rivers", "Grid", "Time", "Open to rebuild"),
  `PCIC VIC-GL, through fwapg` = c(
    "A physically based model of snow, glaciers, soil and runoff, calibrated over 1985–2005",
    sprintf("Peace, Fraser and Columbia (%d watershed groups); fwapg's table leaves out rivers of order 8 and up",
            prov$fwapg_groups),
    sprintf("%s mean absolute error; %s within ±%d%%", pct1(mae_fw), pct(100 * in20_fw), err_cut),
    "Not in fwapg's table; wet can rebuild them from PCIC's output (`scripts/mad_basin.R`, run for the Fraser)",
    sprintf("0.0625°, %s", fmt_cell(km_pcic, 0)),
    "Daily, 1945–2012; twelve climate projections to 2099",
    "Output is public; the model runs are PCIC's"),
  `wet's open water balance` = c(
    "Precipitation less evapotranspiration per grid cell, adjusted by hydrologic zone against HYDAT gauges",
    "All of BC: every segment with a watershed",
    sprintf("%s mean absolute error (held out); %s within ±%d%%", pct1(mae_wb), pct(100 * in20_wb), err_cut),
    sprintf("%s mean absolute error at the %d gauges in PCIC's area where fwapg has no value, mostly on these",
            pct1(mean(abs(dropped$err_pct))), nrow(dropped)),
    sprintf("30″, %s", fmt_cell(km_wb, 1)),
    "1981–2010 normals, annual and monthly",
    "Code ships with wet; its inputs are public and fetched by its scripts"),
  check.names = FALSE
)
err_lim <- c(-100, 200)
n_off <- sum(both$err_pct > err_lim[2] | both$fw_err > err_lim[2])
cap_cmp <- sprintf(paste(
  "Error at each of the %d calibration gauges both estimates cover: fwapg's across, the water balance's",
  "(held out) up. The shaded square holds the gauges both put within ±%d%%. Between the dashed diagonals,",
  "the left and right wedges hold gauges where the balance is closer, the top and bottom wedges those",
  "where fwapg is.%s"
), n_both, err_cut,
if (n_off) sprintf(" %d gauge%s beyond +%d%% %s drawn at the edge.", n_off, if (n_off > 1) "s" else "",
                   err_lim[2], if (n_off > 1) "are" else "is") else "")
```

Two estimates of mean annual discharge are available per segment, built
different ways. They suit different jobs.

```{r models}
knitr::kable(models, align = "lll", row.names = FALSE,
             caption = paste(
               "The two estimates. The balance is scored at gauges held out of its fit, and that score",
               "is mildly optimistic (see Limits); the gauges PCIC calibrated to are not listed with its",
               "output. The two fits are not strictly like for like."
             ))
```

Which to use depends on the job:

- Inside the Peace, Fraser and Columbia, where fwapg has a value: PCIC's
  model, which fits gauges more closely there.
- Outside those basins, on the rivers fwapg leaves out, or for one consistent
  layer across BC: the open water balance.
- Where both exist and the number matters: both, checked against the nearest
  gauge, as the Salmon River section below does.

```{r compare, fig.width = 5.5, fig.height = 5.5, out.width = "80%", fig.cap = cap_cmp}
sq <- data.frame(xmin = -err_cut, xmax = err_cut, ymin = -err_cut, ymax = err_cut)
both_plot <- both
both_plot$fw_err <- pmin(both_plot$fw_err, err_lim[2])
both_plot$err_pct <- pmin(both_plot$err_pct, err_lim[2])
ggplot(both_plot, aes(fw_err, err_pct)) +
  geom_rect(data = sq, aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax), inherit.aes = FALSE,
            fill = pal_prov$values[["zone"]], alpha = 0.6) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = pal_par$values[["one_to_one"]],
              linewidth = 0.3) +
  geom_abline(slope = -1, intercept = 0, linetype = "dashed", colour = pal_par$values[["one_to_one"]],
              linewidth = 0.3) +
  geom_point(aes(colour = nesting), size = 1.6, alpha = 0.8) +
  scale_colour_manual(values = pal_skill$values[c("headwater", "nested")],
                      labels = pal_skill$labels[c("headwater", "nested")], name = NULL) +
  scale_x_continuous(limits = err_lim, labels = function(x) sub("^\\+0%$", "0%", signed(x))) +
  scale_y_continuous(limits = err_lim, labels = function(x) sub("^\\+0%$", "0%", signed(x))) +
  annotate("text", x = 0.75 * err_lim[2], y = 0.1 * err_lim[2], label = "balance closer", size = 3,
           colour = pal_par$values[["one_to_one"]]) +
  annotate("text", x = 0.1 * err_lim[2], y = 0.9 * err_lim[2], label = "fwapg closer", size = 3,
           colour = pal_par$values[["one_to_one"]]) +
  coord_equal() +
  labs(x = "fwapg error", y = "Water balance error (held out)")
```

## How close it is, province-wide

```{r skill-data}
mae_of <- function(i) mean(abs(sk$err_pct[i]))
mae_all <- mae_of(TRUE)
mae_head <- mae_of(sk$nesting == "headwater")
mae_nest <- mae_of(sk$nesting == "nested")
in20 <- mean(abs(sk$err_pct) <= err_cut)
small_km2 <- 100
large_km2 <- 1000
mae_small <- mae_of(sk$area_km2 < small_km2)
mae_large <- mae_of(sk$area_km2 >= large_km2)
zone_mae <- sort(tapply(abs(sk$err_pct), sk$zone, mean))
worst <- names(zone_mae)[length(zone_mae)]
worst_name <- to_title(map$zones$name[map$zones$zone == worst])
n_worst <- sum(sk$zone == worst)
# the prose: nested gauges score better than headwater, small basins worse than
# large, and the worst zone is a semi-arid interior plateau (15, 17, 23 or 24:
# research/water_balance_method.md) that stands out at twice the average
stopifnot(mae_nest < mae_head, mae_small > mae_large, sum(zone_mae > 2 * mae_all) == 1,
          worst %in% c("15", "17", "23", "24"), length(worst_name) == 1)
sk$class <- err_class(sk$err_pct)
cap_prov <- sprintf(paste(
  "The %d calibration gauges, filled by the water balance's held-out error at each. PCIC's coverage,",
  "where fwapg also has values, is outlined; zone %s, the weakest, is shaded; the two groups mapped",
  "below are in black."
), nrow(sk), worst)
cap_hist <- sprintf(paste(
  "Held-out error at the %d calibration gauges, in classes of %d%%, coloured by the error scale.",
  "Gauges beyond +200%% are counted in the last bar."
), nrow(sk), 10)
```

Out of sample, the balance's annual runoff at a gauge is off by
`r pct(mae_all)` on average, and `r pct(100 * in20)` of gauges fall within
±`r err_cut`%. It does better on larger rivers. Nested gauges average
`r pct(mae_nest)` against `r pct(mae_head)` for headwater ones, and basins
over `r num(large_km2)` km² average `r pct(mae_large)` against
`r pct(mae_small)` for those under `r small_km2` km². One zone stands out:
zone `r worst`, the `r worst_name`, a semi-arid interior plateau, where the
`r n_worst` gauges average `r pct(zone_mae[[worst]])`.

```{r map-province, fig.width = 7, fig.height = 6.6, fig.cap = cap_prov}
gpts <- as_pts(sk)
zone_w <- map$zones[map$zones$zone == worst, ]
bb_bc <- gq::gq_bbox_aspect(sf::st_bbox(bc), asp = 7 / 5.6)
lab_g <- sf::st_centroid(map$groups)
lab_g$label <- group_name[lab_g$watershed_group_code]
xy_g <- sf::st_coordinates(lab_g)
dx_bc <- bb_bc[["xmax"]] - bb_bc[["xmin"]]
dy_bc <- bb_bc[["ymax"]] - bb_bc[["ymin"]]
sb_km <- max(gq::gq_scale_breaks(bb_bc))
sb <- data.frame(x = bb_bc[["xmin"]] + 0.05 * dx_bc, y = bb_bc[["ymin"]] + 0.05 * dy_bc)
sb$xend <- sb$x + sb_km * 1000
ggplot() +
  geom_sf(data = bc, fill = "white", colour = pal_map$values[["outline"]], linewidth = 0.25) +
  geom_sf(data = zone_w, aes(alpha = "zone"), fill = pal_prov$values[["zone"]], colour = NA) +
  geom_sf(data = map$coverage, aes(linetype = "coverage"), fill = NA, colour = pal_prov$values[["coverage"]],
          linewidth = 0.45) +
  geom_sf(data = map$groups, fill = NA, colour = pal_prov$values[["group"]], linewidth = 0.6) +
  geom_label(data = data.frame(x = xy_g[, 1] + 0.03 * dx_bc, y = xy_g[, 2], label = lab_g$label),
             aes(x, y, label = label), hjust = 0, size = 2.8, linewidth = 0, fill = "white", alpha = 0.6,
             label.padding = unit(1.2, "pt")) +
  geom_sf(data = gpts, aes(fill = class), shape = 21, colour = "black", size = 1.7, stroke = 0.25) +
  geom_segment(data = sb, aes(x = x, xend = xend, y = y, yend = y), linewidth = 1) +
  geom_text(data = sb, aes(x = (x + xend) / 2, y = y, label = paste(sb_km, "km")), vjust = -0.7, size = 3) +
  err_scale("Held-out error") +
  scale_alpha_manual(values = c(zone = 1), labels = sprintf("Zone %s, %s", worst, worst_name), name = NULL) +
  scale_linetype_manual(values = c(coverage = "dashed"), labels = pal_prov$labels[["coverage"]], name = NULL) +
  coord_sf(xlim = bb_bc[c("xmin", "xmax")], ylim = bb_bc[c("ymin", "ymax")], expand = FALSE, datum = NA) +
  guides(fill = guide_legend(order = 1, ncol = 2), alpha = guide_legend(order = 2),
         linetype = guide_legend(order = 3)) +
  theme_void() +
  theme(legend.position = "bottom", legend.box = "vertical", legend.text = element_text(size = 8),
        panel.border = element_rect(fill = NA, colour = pal_map$values[["outline"]], linewidth = 0.3))
```

```{r hist, fig.height = 2.8, fig.cap = cap_hist}
sk_h <- sk
sk_h$err_plot <- pmin(sk_h$err_pct, 205)
ggplot(sk_h, aes(err_plot, fill = class)) +
  geom_histogram(breaks = seq(-100, 210, by = 10), colour = "white", linewidth = 0.2) +
  geom_vline(xintercept = c(-err_cut, err_cut), linetype = "dashed", colour = pal_par$values[["one_to_one"]],
             linewidth = 0.3) +
  err_scale("Held-out error") +
  scale_x_continuous(breaks = c(-100, -50, -20, 0, 20, 50, 100, 150, 200),
                     labels = function(x) sub("^\\+0%$", "0%", signed(x))) +
  labs(x = "Held-out error in mean annual runoff", y = "Gauges") +
  theme(legend.position = "none")
```

## Salmon River: two estimates side by side

```{r map-data}
seg <- merge(map$segments, v$segments, by = c("linear_feature_id", "watershed_group_code"))
stopifnot(nrow(seg) == nrow(map$segments))
n_no_value <- sum(is.na(seg$mad_wb_m3s))
# the limits bullet says neither product has these
stopifnot(all(is.na(seg$mad_pcic_m3s[is.na(seg$mad_wb_m3s)])))
order_no_value <- range(seg$stream_order[is.na(seg$mad_wb_m3s)])
n_total <- table(seg$watershed_group_code)
seg <- seg[!is.na(seg$mad_wb_m3s), ]

# the scale bar's place, and the box it and its text take, which labels avoid
scale_bar_at <- function(bb, at = "right") {
  dx <- bb[["xmax"]] - bb[["xmin"]]
  dy <- bb[["ymax"]] - bb[["ymin"]]
  km <- max(gq::gq_scale_breaks(bb))
  x0 <- if (at == "right") bb[["xmax"]] - 0.05 * dx - km * 1000 else bb[["xmin"]] + 0.05 * dx
  sb <- data.frame(x = x0, y = bb[["ymin"]] + 0.04 * dy, km = km)
  sb$xend <- sb$x + km * 1000
  sb$x0 <- sb$x
  sb$x1 <- sb$xend
  sb$y0 <- sb$y - 0.01 * dy
  sb$y1 <- sb$y + 0.06 * dy
  sb
}
scale_bar <- function(bb, at = "right") {
  sb <- scale_bar_at(bb, at)
  list(geom_segment(data = sb, aes(x = x, xend = xend, y = y, yend = y), linewidth = 1),
       geom_text(data = sb, aes(x = (x + xend) / 2, y = y, label = paste(km, "km")), vjust = -0.7, size = 3))
}
mad_scales <- list(
  scale_colour_manual(values = pal_mad$values, labels = pal_mad$labels, name = "Mean annual\ndischarge (m³/s)",
                      drop = FALSE),
  scale_linewidth_manual(values = stats::setNames(mad_width, names(pal_mad$values)), labels = pal_mad$labels,
                         name = "Mean annual\ndischarge (m³/s)", drop = FALSE)
)
# labels for towns and gauges, placed greedily: gauges first, each beside its
# point (right, left, above, below) where it overlaps no point, no label placed
# before it and no frame edge; failing that, where it overlaps least. Box sizes
# are estimated from the text size and the panel's size in inches
map_labels <- function(towns, gauges, bb, width_in, height_in, avoid = NULL) {
  in_bb <- function(xy) {
    xy[, 1] > bb[["xmin"]] & xy[, 1] < bb[["xmax"]] & xy[, 2] > bb[["ymin"]] & xy[, 2] < bb[["ymax"]]
  }
  tx <- sf::st_coordinates(towns)
  towns <- towns[in_bb(tx), ]
  tx <- tx[in_bb(tx), , drop = FALSE]
  gx <- sf::st_coordinates(gauges)
  ux <- (bb[["xmax"]] - bb[["xmin"]]) / width_in    # map units per inch
  uy <- (bb[["ymax"]] - bb[["ymin"]]) / height_in
  item <- data.frame(x = c(gx[, 1], tx[, 1]), y = c(gx[, 2], tx[, 2]),
                     label = c(gauges$station_number, towns$name),
                     size = rep(c(2.6, 2.8), c(nrow(gx), nrow(tx))),
                     town = rep(c(FALSE, TRUE), c(nrow(gx), nrow(tx))))
  pt <- item$size * 72.27 / 25.4
  item$w <- nchar(item$label) * pt * ifelse(item$town, 0.66, 0.6) / 72 * ux
  item$h <- pt * 1.5 / 72 * uy
  r <- 0.06 * ux   # half a symbol, as an obstacle
  boxes <- data.frame(x0 = item$x - r, x1 = item$x + r, y0 = item$y - r, y1 = item$y + r)
  if (!is.null(avoid)) boxes <- rbind(boxes, avoid[c("x0", "x1", "y0", "y1")])
  overlap <- function(b, others) {
    ox <- pmax(0, pmin(b$x1, others$x1) - pmax(b$x0, others$x0))
    oy <- pmax(0, pmin(b$y1, others$y1) - pmax(b$y0, others$y0))
    sum(ox * oy)
  }
  gap <- 0.07 * ux
  out <- vector("list", nrow(item))
  for (i in seq_len(nrow(item))) {
    w <- item$w[i]
    h <- item$h[i]
    x <- item$x[i]
    y <- item$y[i]
    cand <- data.frame(hjust = c(0, 1, 0.5, 0.5), vjust = c(0.5, 0.5, 0, 1),
                       lx = c(x + gap, x - gap, x, x), ly = c(y, y, y + gap, y - gap))
    cand$x0 <- cand$lx - cand$hjust * w
    cand$x1 <- cand$x0 + w
    cand$y0 <- cand$ly - cand$vjust * h
    cand$y1 <- cand$y0 + h
    cost <- vapply(seq_len(nrow(cand)), function(k) {
      b <- cand[k, ]
      edge <- (b$x0 < bb[["xmin"]]) + (b$x1 > bb[["xmax"]]) + (b$y0 < bb[["ymin"]]) + (b$y1 > bb[["ymax"]])
      overlap(b, boxes[-i, ]) + edge * w * h
    }, 0)
    k <- which.min(cost)
    out[[i]] <- cand[k, c("lx", "ly", "hjust", "vjust")]
    boxes <- rbind(boxes, cand[k, c("x0", "x1", "y0", "y1")])
  }
  lab <- cbind(item[c("label", "town")], do.call(rbind, out))
  list(geom_point(data = data.frame(x = tx[, 1], y = tx[, 2]), aes(x, y), size = 1.1,
                  colour = town_sty$label$color),
       geom_label(data = lab[lab$town, ], aes(lx, ly, label = label, hjust = hjust, vjust = vjust),
                  colour = town_sty$label$color, fontface = "bold", size = 2.8, linewidth = 0, fill = "white",
                  alpha = 0.8, label.padding = unit(1.2, "pt")),
       geom_label(data = lab[!lab$town, ], aes(lx, ly, label = label, hjust = hjust, vjust = vjust),
                  size = 2.6, linewidth = 0, fill = "white", alpha = 0.85, label.padding = unit(1, "pt")))
}
map_theme <- theme_void() +
  theme(panel.border = element_rect(fill = NA, colour = pal_map$values[["outline"]], linewidth = 0.3))

# SALR: both products, and the gauges that arbitrate between them
ss <- v$segments[v$segments$watershed_group_code == "SALR" & !is.na(v$segments$mad_wb_m3s), ]
ratio_salr <- stats::median(ss$mad_wb_m3s / ss$mad_pcic_m3s)
ga <- v$gauges_salr
ga$wb_err <- 100 * (ga$wb_mm / ga$obs_mm - 1)
ga$fw_err <- 100 * (ga$fwapg_mm / ga$obs_mm - 1)
outlet <- ga[ga$holds_salr, ]
# the balance's share of the gap at each gauge, in log terms like the ratio:
# log(wb / fwapg) = log(wb / obs) + log(obs / fwapg)
ga$wb_share <- log(ga$wb_mm / ga$obs_mm) / log(ga$wb_mm / ga$fwapg_mm)
ga$close <- abs(ga$fw_err) <= 5   # a column, so it follows the reorder below
close <- ga$close
# the prose: the balance reads high at every gauge; at the small basins fwapg is
# within 5% and the gap is nearly all the balance's; at the large ones, the
# outlet among them, fwapg is low and the two share it
stopifnot(ratio_salr > 1, all(ga$wb_err > 0), nrow(outlet) == 1,
          any(close), any(!close), max(ga$area_km2[close]) < min(ga$area_km2[!close]),
          all(ga$wb_share[close] > 0.9), all(ga$fw_err[!close] < -5),
          all(ga$wb_share[!close] > 0.25 & ga$wb_share[!close] < 0.75), !close[ga$holds_salr])
ga <- ga[order(!ga$holds_salr, ga$area_km2), ]
close <- ga$close
named <- c(named, ga$station_number)
gauge_tab <- data.frame(
  Gauge = sprintf("%s %s", ga$station_number, to_title(ga$station_name)),
  `Area (km²)` = num(ga$area_km2),
  `Observed (mm)` = num(ga$obs_mm),
  `Water balance (held out)` = signed(ga$wb_err),
  fwapg = signed(ga$fw_err),
  check.names = FALSE
)

# the Salmon River frame: the group, every gauge in the table and every town a
# gauge is named for ("near Prince George"), inside the context box the map
# layers were cut to
salr <- map$groups[map$groups$watershed_group_code == "SALR", ]
ga_pts <- as_pts(ga)
salr_w <- 3.4
salr_h <- 4.4
pl_salr <- map$places[map$places$watershed_group_code == "SALR", ]
pl_near <- pl_salr[vapply(toupper(pl_salr$name), function(n) any(grepl(n, ga$station_name, fixed = TRUE)), NA), ]
stopifnot(nrow(pl_near) > 0)
bb_salr <- sf::st_bbox(sf::st_buffer(sf::st_union(c(sf::st_geometry(salr), sf::st_geometry(ga_pts),
                                                    sf::st_geometry(pl_near))), 8000))
bb_salr <- gq::gq_bbox_aspect(bb_salr, asp = salr_w / salr_h)
ctx <- sf::st_bbox(map$context)
stopifnot(bb_salr[["xmin"]] >= ctx[["xmin"]], bb_salr[["xmax"]] <= ctx[["xmax"]],
          bb_salr[["ymin"]] >= ctx[["ymin"]], bb_salr[["ymax"]] <= ctx[["ymax"]])
drawn <- c(drawn, in_frame(ga_pts, bb_salr))

products <- c(fwapg = "fwapg (PCIC)", wb = "Open water balance")
s2 <- seg[seg$watershed_group_code == "SALR", ]
s2 <- rbind(transform(s2, product = products[["fwapg"]], class = mad_class(s2$mad_pcic_m3s)),
            transform(s2, product = products[["wb"]], class = mad_class(s2$mad_wb_m3s)))
s2$product <- factor(s2$product, products)
s2 <- s2[order(s2$class), ]   # large rivers drawn last, on top
g2 <- rbind(transform(ga_pts, product = products[["fwapg"]], class = err_class(ga_pts$fw_err)),
            transform(ga_pts, product = products[["wb"]], class = err_class(ga_pts$wb_err)))
g2$product <- factor(g2$product, products)
lakes_salr <- rbind(map$lakes[map$lakes$watershed_group_code == "SALR", c("gnis_name", "area_ha")],
                    map$context_lakes[c("gnis_name", "area_ha")])
cap_salr <- sprintf(paste(
  "Mean annual discharge on %s of the %s %s segments of order %d or more, from fwapg (left) and the",
  "open water balance (right), with the gauges of the table filled by each product's error. Rivers",
  "outside the group are grey."
), num(nrow(seg[seg$watershed_group_code == "SALR", ])), num(n_total[["SALR"]]), group_name[["SALR"]],
prov$min_order)
```

Both products cover the `r group_name[["SALR"]]`, a headwater group of the
Fraser (`r num(group_km2[["SALR"]])` km²), and they differ there. The open
water balance gives a median `r format(round(ratio_salr, 2), nsmall = 2)`
times fwapg's value per segment. The `r nrow(ga)` gauges in and around the
group show which is closer where.

The balance runs high at all of them, by `r signed(min(ga$wb_err))` to
`r signed(max(ga$wb_err))`. At the `r sum(close)` smallest basins, fwapg is
within 5%, so the gap there is the balance's. At the `r sum(!close)` larger
ones, fwapg runs `r signed(max(ga$fw_err[!close]))` to
`r signed(min(ga$fw_err[!close]))`, and the two share the gap. At
`r outlet$station_number`, the gauge that drains the whole group, they
bracket the observation: `r signed(outlet$wb_err)` against
`r signed(outlet$fw_err)`.

```{r gauges-salr}
knitr::kable(gauge_tab, align = "lrrrr", row.names = FALSE,
             caption = sprintf(paste(
               "Calibration gauges within %d km of the %s group, and its outlet gauge (first row):",
               "mean annual runoff observed in 1981–2010, and each product's error against it."
             ), prov$near_km, group_name[["SALR"]]))
```

```{r map-salr, fig.width = 7, fig.height = 5.6, fig.cap = cap_salr}
ggplot() +
  geom_sf(data = salr, fill = pal_map$values[["group"]], colour = pal_map$values[["outline"]], linewidth = 0.4) +
  geom_sf(data = lakes_salr, fill = lake_sty$fill$color, colour = lake_sty$stroke$color, linewidth = 0.15) +
  geom_sf(data = map$context_streams, aes(alpha = "river"), colour = pal_ctx$values[["river"]], linewidth = 0.45) +
  scale_alpha_manual(values = c(river = 1), labels = pal_ctx$labels[["river"]], name = NULL) +
  geom_sf(data = s2, aes(colour = class, linewidth = class)) +
  mad_scales +
  map_labels(map$places[map$places$watershed_group_code == "SALR", ], ga_pts, bb_salr, salr_w, salr_h,
             avoid = scale_bar_at(bb_salr, "left")) +
  geom_sf(data = g2, aes(fill = class), shape = 21, colour = "black", size = 2.4, stroke = 0.4,
          show.legend = c(fill = TRUE, colour = FALSE, linewidth = FALSE, alpha = FALSE)) +
  err_scale() +
  scale_bar(bb_salr, "left") +   # Prince George sits bottom right
  facet_wrap(~product) +
  coord_sf(xlim = bb_salr[c("xmin", "xmax")], ylim = bb_salr[c("ymin", "ymax")], expand = FALSE, datum = NA) +
  guides(colour = guide_legend(order = 1, ncol = 2), linewidth = guide_legend(order = 1, ncol = 2),
         fill = guide_legend(order = 2, ncol = 1, override.aes = list(shape = 21, size = 2.6, colour = "black")),
         alpha = guide_legend(order = 3)) +
  map_theme +
  theme(legend.position = "bottom", strip.text = element_text(size = 10, face = "bold"),
        legend.text = element_text(size = 8), legend.title = element_text(size = 8.5))
```

## Bulkley River: the water balance alone

```{r bulk-data}
gb <- sk[sk$watershed_group_code == "BULK", ]
gb <- gb[order(gb$area_km2, decreasing = TRUE), ]
gb_pts <- as_pts(gb)
gb_pts$class <- err_class(gb_pts$err_pct)
nb <- nrow(gb)
bulk_mae <- mean(abs(gb$err_pct))
bulk_zones <- sort(unique(gb$zone))
stopifnot(nb > 0, prov$fwapg_discharge_rows[["BULK"]] == 0)
named <- c(named, gb$station_number)
bulk_tab <- data.frame(
  Gauge = sprintf("%s %s", gb$station_number, to_title(gb$station_name)),
  `Area (km²)` = num(gb$area_km2),
  `Observed (mm)` = num(gb$obs),
  `Water balance (mm, held out)` = num(gb$mod),
  Error = signed(gb$err_pct),
  check.names = FALSE
)

bulk <- map$groups[map$groups$watershed_group_code == "BULK", ]
bulk_w <- 4.6
bulk_h <- 6.4
bb_bulk <- gq::gq_bbox_aspect(sf::st_bbox(sf::st_buffer(bulk, 4000)), asp = bulk_w / bulk_h)
drawn <- c(drawn, in_frame(gb_pts, bb_bulk))
sb_bulk <- seg[seg$watershed_group_code == "BULK", ]
sb_bulk$class <- mad_class(sb_bulk$mad_wb_m3s)
sb_bulk <- sb_bulk[order(sb_bulk$class), ]

# the worked example: the largest gauge, traced from its segment's value
ex <- gb[1, ]
ex_seg <- v$segments[v$segments$linear_feature_id == ex$linear_feature_id, ]
stopifnot(nrow(ex_seg) == 1, !is.na(ex_seg$mad_wb_m3s))   # the gauge's segment is in the bundle
ex_obs <- wet_mm_to_m3s(ex$obs, ex$area_km2 * 1e6)
ex_ho <- wet_mm_to_m3s(ex$mod, ex$area_km2 * 1e6)
ex_map_err <- 100 * (ex_seg$mad_wb_m3s / ex_obs - 1)
# the prose: the map's value, fitted with this gauge in, is nearer the gauge
# than the held-out prediction
stopifnot(round(abs(ex_map_err)) < round(abs(ex$err_pct)))   # as printed, whole percents

key <- ggplot() +
  geom_sf(data = bc, fill = "white", colour = pal_map$values[["outline"]], linewidth = 0.2) +
  geom_sf(data = bulk, fill = pal_map$values[["key"]], colour = NA) +
  geom_sf(data = sf::st_centroid(sf::st_geometry(bulk)), shape = 1, size = 6, stroke = 0.6,
          colour = pal_map$values[["key"]]) +
  theme_void() +
  theme(panel.background = element_rect(fill = "white", colour = pal_map$values[["outline"]], linewidth = 0.3))
cap_bulk <- sprintf(paste(
  "Mean annual discharge on %s of the %s %s segments of order %d or more, from the open water",
  "balance, and the %d calibration gauges in the group, filled by the balance's held-out error.",
  "The keymap marks the group in BC."
), num(nrow(sb_bulk)), num(n_total[["BULK"]]), group_name[["BULK"]], prov$min_order, nb)
```

fwapg has no values in the `r group_name[["BULK"]]`, in the Skeena
(`r num(group_km2[["BULK"]])` km²), which lies outside PCIC's coverage. The
open water balance is the only estimate there. Its `r nb` gauges average
`r pct(bulk_mae)` held-out error, from `r signed(min(gb$err_pct))` to
`r signed(max(gb$err_pct))`, in zones `r and_text(bulk_zones)`.

The map's segment values come from the fit that includes every gauge, while
the gauges are filled by the held-out error, so a gauge can look worse than
the segment it sits on. Take `r ex$station_number`, `r to_title(ex$station_name)`
(`r num(ex$area_km2)` km²). It measured `r num(ex$obs)` mm a year, or
`r num(ex_obs, 1)` m³/s over its basin. Predicted with its sub-sub-drainage
held out, the balance gives `r num(ex$mod)` mm, or `r num(ex_ho, 1)` m³/s
(`r signed(ex$err_pct)`). The segment it sits on reads `r num(ex_seg$mad_wb_m3s, 1)`
m³/s on the map (`r signed(ex_map_err)`), closer, because that fit saw this
gauge.

```{r map-bulk, fig.width = bulk_w + 1.8, fig.height = bulk_h, fig.cap = cap_bulk}
key_w <- 0.34 * (bb_bulk[["xmax"]] - bb_bulk[["xmin"]])
key_h <- key_w * 1.05
dxb <- bb_bulk[["xmax"]] - bb_bulk[["xmin"]]
dyb <- bb_bulk[["ymax"]] - bb_bulk[["ymin"]]
ggplot() +
  geom_sf(data = bulk, fill = pal_map$values[["group"]], colour = pal_map$values[["outline"]], linewidth = 0.4) +
  geom_sf(data = map$lakes[map$lakes$watershed_group_code == "BULK", ], fill = lake_sty$fill$color,
          colour = lake_sty$stroke$color, linewidth = 0.15) +
  geom_sf(data = sb_bulk, aes(colour = class, linewidth = class)) +
  mad_scales +
  map_labels(map$places[map$places$watershed_group_code == "BULK", ], gb_pts, bb_bulk, bulk_w, bulk_h,
             avoid = scale_bar_at(bb_bulk)) +
  geom_sf(data = gb_pts, aes(fill = class), shape = 21, colour = "black", size = 2.6, stroke = 0.4,
          show.legend = c(fill = TRUE, colour = FALSE, linewidth = FALSE, alpha = FALSE)) +
  err_scale("Held-out error") +
  scale_bar(bb_bulk) +
  annotation_custom(ggplotGrob(key), xmin = bb_bulk[["xmax"]] - key_w - 0.02 * dxb,
                    xmax = bb_bulk[["xmax"]] - 0.02 * dxb,
                    ymin = bb_bulk[["ymax"]] - key_h - 0.02 * dyb, ymax = bb_bulk[["ymax"]] - 0.02 * dyb) +
  coord_sf(xlim = bb_bulk[c("xmin", "xmax")], ylim = bb_bulk[c("ymin", "ymax")], expand = FALSE, datum = NA) +
  map_theme +
  theme(legend.position = "right", legend.key.width = unit(14, "pt"))
```

```{r bulk-table}
knitr::kable(bulk_tab, align = "lrrrr", row.names = FALSE,
             caption = sprintf(paste(
               "The %d calibration gauges in the %s group: mean annual runoff observed in 1981–2010,",
               "the water balance's held-out prediction, and its error."
             ), nb, group_name[["BULK"]]))
```

## Limits

```{r limits}
stale <- prov$upstream_area_100
# the bullet: stale polygons exist in the Fraser, and none moves a SALR segment
stopifnot(stale[["stale"]] > 0, prov$salr_stale_segments == 0)
small_share <- mean(v$sampling$upstream_area_km2 < small_km2)
```

In order of how much they affect a user:

- **Coverage.** Outside the Peace, Fraser and Columbia, the open water
  balance is the only estimate.
- **Dry zones.** In zone `r worst` and the other semi-arid interior zones,
  runoff is a small remainder of precipitation, so a small error in
  evapotranspiration is a large one in runoff. The zone adjustment
  passes its gate only under a variant offered after the pre-set one failed,
  so all the skill figures here are mildly optimistic
  (`research/water_balance_method.md`).
- **Small streams.** Gauged basins under `r small_km2` km² average
  `r pct(mae_small)` error, and `r pct(100 * small_share)` of the
  `r group_name[["SALR"]]`'s watersheds with a stream are that small.
- **Segments without a watershed.** In the two groups mapped here,
  `r n_no_value` segments of order `r order_no_value[1]` to
  `r order_no_value[2]` have no watershed polygon in the Freshwater Atlas
  lookup, so neither product gives them a value. They are
  left off the maps, which leaves short gaps in some streams.

## How it is built

```{r parity-data}
p <- v$parity
stopifnot(p$n_segments == prov$fwapg_discharge_rows[["SALR"]], p$n_match5 == p$n_segments)
```

wet rebuilds fwapg's numbers from PCIC's output with its own code, and on
every one of the `r num(p$n_segments)` `r group_name[["SALR"]]` segments
fwapg gives a value, it matches fwapg to the five decimals fwapg stores.

```{r sampling-data}
sm <- v$sampling
sm$change <- 100 * (sm$mm_area / sm$mm_centroid - 1)
area_cut <- c(0, 10, 100, Inf)
sm$size <- cut(sm$upstream_area_km2, area_cut)
moved <- tapply(abs(sm$change) > 5, sm$size, mean)
q99 <- tapply(abs(sm$change), sm$size, stats::quantile, 0.99)
# the prose: small watersheds move, large ones barely
stopifnot(moved[[1]] > moved[[3]], q99[[1]] > 5 * q99[[3]])
cell_ns <- km_pcic[["ns"]]   # the table's PCIC cell, at the same latitude
cell_ew <- km_pcic[["ew"]]
cap_samp <- sprintf(paste(
  "Change in mean annual runoff from area-weighted sampling, against centroid sampling, for",
  "the %s %s watersheds that carry a stream. Upstream area is on a logarithmic axis."
), num(nrow(sm)), group_name[["SALR"]])
```

fwapg gives each watershed the value of the PCIC cell under its centroid. A
cell is 0.0625°, about `r round(cell_ns)` by `r round(cell_ew)` km here, so a
small watershed gets one cell's value. wet can instead weight every cell by
the area it covers. Under `r area_cut[2]` km², `r pct(100 * moved[[1]])` of
watersheds move by more than 5%, by up to `r pct(q99[[1]])` at the 99th
percentile. Above `r area_cut[3]` km², the 99th percentile is
`r format(round(q99[[3]], 1), nsmall = 1)`%, because a large basin averages
many cells either way.

```{r sampling, fig.width = 5.5, fig.height = 3, out.width = "75%", fig.cap = cap_samp}
ggplot(sm, aes(upstream_area_km2, change)) +
  geom_hline(yintercept = 0, colour = pal_par$values[["one_to_one"]], linewidth = 0.3) +
  geom_point(colour = pal_par$values[["segment"]], size = 0.5, alpha = 0.4) +
  scale_x_log10(labels = function(x) format(x, big.mark = ",", scientific = FALSE, drop0trailing = TRUE)) +
  labs(x = "Upstream area (km²)", y = "Change (%)")
```

fwapg stores each watershed's upstream area from an earlier build of the
Freshwater Atlas. In the Fraser, `r num(stale[["stale"]])` of
`r num(stale[["polygons"]])` polygons carry a stored area that differs from the
area accumulated from the Atlas today. None of them changes a `r group_name[["SALR"]]` segment, so the match
above does not test it. wet uses the accumulated area except when it
reproduces fwapg.

The PCIC path for one headwater group, as `scripts/mad_parity.R` runs it,
needs fwapg and the PCIC server. The open water balance is a pipeline of
scripts (`scripts/wb_inputs.R` to `scripts/wb_output.R`; see the package's
`CLAUDE.md`).

```{r recipe, echo = TRUE, eval = FALSE}
library(wet)
conn <- DBI::dbConnect(RPostgres::Postgres(), dbname = "fwapg")

# PCIC VIC-GL runoff and baseflow over the group's extent, 1981-2010
bbox <- c(-124.12, 54.22, -123.12, 55.11)
runoff <- terra::rast(wet_pcic_fetch("RUNOFF", bbox, "1981-01-01", "2010-12-31"))
baseflow <- terra::rast(wet_pcic_fetch("BASEFLOW", bbox, "1981-01-01", "2010-12-31"))
mad_cell <- wet_runoff_annual(runoff) + wet_runoff_annual(baseflow)   # mm/yr per cell

# every watershed polygon in the Fraser; those inside the extent get a cell value
ws <- wet_ws_fetch(conn, "100")
in_bb <- ws$lon >= bbox[1] & ws$lon <= bbox[3] & ws$lat >= bbox[2] & ws$lat <= bbox[4]
cells <- wet_ws_sample(mad_cell, ws[in_bb, c("watershed_feature_id", "lon", "lat")], "centroid")
up <- wet_upstream_mean(ws, cells, "total", irregular_pairs = wet_upstream_irregular(conn, "100"))
up$mad_m3s <- wet_mm_to_m3s(up$value, up$upstream_area_m2)
```

**Cached inputs.** This article runs on saved outputs: PCIC VIC-GL through
`scripts/mad_parity.R`; the open water balance from province run
`r prov$province_run` (annual AET `r prov$aet`, HYDAT release
`r substr(prov$hydat_release, 1, 10)`); fwapg's values at the gauges; built
`r paste(as.integer(format(prov$built, "%d")), format(prov$built, "%B %Y"))` at commit `r prov$wet_commit`. Regenerated by
`data-raw/segment_vignette_data.R`. Map layers from the Freshwater Atlas
(fwapg), BC Geographic Names and the BC Hydrologic Zones, by
`data-raw/segment_vignette_map.R`.

```{r named-on-map}
# every station named in the prose or a table is inside a map's frame: the
# tables add theirs above, and any station number typed into the prose is added here
f <- knitr::current_input()
if (!is.null(f)) {
  x <- readLines(f, warn = FALSE)
  named <- c(named, unlist(regmatches(x, gregexpr("\\b0[0-9][A-Z]{2}[0-9]{3}\\b", x))))
}
named <- unique(c(named, outlet$station_number, ex$station_number))
missing <- setdiff(named, drawn)
if (length(missing)) stop("named but not on a map: ", paste(missing, collapse = ", "))
```

```{r wordcount, include = FALSE}
if (!is.null(f)) {
  y <- utils::tail(x, -utils::head(which(x == "---"), 2)[2])   # drop the YAML header
  fence <- cumsum(grepl("^```", y)) %% 2 == 1
  y <- y[!fence & !grepl("^```", y)]
  y <- gsub("`r [^`]*`", "", y)          # inline code is not body prose
  y <- y[!grepl("^\\s*$", y)]
  n <- sum(lengths(strsplit(trimws(y), "\\s+")))
  message("body prose: ", n, " words")
  cap <- 1600
  if (n > cap) stop("body prose is ", n, " words; the cap is ", cap)
}
```
