---
title: "Flow in Species Windows at Two Stations"
output: bookdown::html_vignette2
vignette: >
  %\VignetteIndexEntry{Flow in Species Windows at Two Stations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.width = 7,
  out.width = "100%",
  dpi = 150
)
```

```{r load}
library(wet)
library(ggplot2)

vd <- function(f) system.file("vignette-data", f, package = "wet")
daily <- readRDS(vd("station_daily.rds"))
prov <- attr(daily, "provenance")
lh <- utils::read.csv(vd("life_history_bulk.csv"), stringsAsFactors = FALSE, na.strings = "")
windows <- data.frame(window = paste(lh$species_code, sub("_", " ", lh$life_stage)),
                      start = lh$start, end = lh$end)

to_title <- function(x) {
  vapply(strsplit(tolower(x), " "), function(w) {
    big <- !w %in% c("at", "the", "near", "of")
    w[big] <- paste0(toupper(substring(w[big], 1, 1)), substring(w[big], 2))
    paste(w, collapse = " ")
  }, "")
}
station_label <- stats::setNames(
  sprintf("%s (%s)", to_title(prov$stations$station_name), prov$stations$station_number),
  prov$stations$station_number
)
stations <- names(station_label)

baseline <- 1981:2010
min_baseline <- 10
trend_start <- c(1981, 2000)
alpha <- 0.05
base_text <- paste0(min(baseline), "–", max(baseline))
trend_text <- paste("from", trend_start, collapse = " and ")
last_year <- as.integer(format(max(daily$date), "%Y"))
recent <- (last_year - 3):last_year

# colours come from the gq registry, so no hex literal appears in this vignette;
# assert each class set, because a malformed CSV row adds a layer silently
reg <- gq::gq_reg_custom(system.file("cartography", "wet_station.csv", package = "wet"))
classes <- function(layer, expect) {
  x <- gq::gq_tmap_classes(reg$layers[[layer]])
  stopifnot(identical(names(x$values), expect))
  list(values = x$values, labels = stats::setNames(x$labels, names(x$values)))
}
pal_source <- classes("record_source", c("hydat_open", "hydat_ice", "provisional", "realtime"))
pal_hydro <- classes("hydrograph", c("band", "median", "window"))
pal_hydro$labels <- sub("^Baseline", base_text, pal_hydro$labels)
# ordinal classes, so a data refresh needs no registry edit
pal_year <- classes("recent_year", c("latest_3", "latest_2", "latest_1", "latest"))
names(pal_year$values) <- names(pal_year$labels) <- pal_year$labels <- as.character(recent)
pal_dep <- classes("departure_class", c(paste0("d", 1:8), "dropped", "no_baseline"))
pal_trend <- classes("trend_start", as.character(trend_start))

theme_set(theme_minimal(base_size = 10) +
            theme(legend.position = "bottom", panel.grid.minor = element_blank()))
pct <- function(x) paste0(round(abs(x)), "%")
day_text <- function(d) paste(as.integer(format(d, "%d")), format(d, "%B %Y"))
by_station <- function(d) {
  d$station_number <- factor(d$station_number, stations)
  d
}
years_text <- function(y) {
  y <- sort(unique(y))
  runs <- split(y, cumsum(c(1, diff(y) != 1)))
  one <- function(r) if (length(r) > 1) paste0(r[1], "–", r[length(r)]) else as.character(r)
  sub(", ([^,]*)$", " and \\1", paste(vapply(runs, one, ""), collapse = ", "))
}
```

```{r stats}
stats <- wet_window_stats(daily, windows, stats = c("mean", "min7"))

# Provisional winter flow is uncorrected stage under ice, so a window touching
# November-April drops any year with a provisional day (as scripts/station_departure.R)
touches_ice <- vapply(seq_len(nrow(windows)), function(k) {
  a <- as.Date(paste0("2001-", windows$start[k]))
  b <- as.Date(paste0(2001 + (windows$end[k] < windows$start[k]), "-",
                      sub("02-29", "02-28", windows$end[k])))
  any(as.integer(format(seq(a, b, by = "day"), "%m")) %in% c(11:12, 1:4))
}, TRUE)
stats$dropped <- stats$period %in% windows$window[touches_ice] & stats$frac_provisional > 0
kept <- stats[!stats$dropped, ]

# baseline years per station and window; under min_baseline gets no departure
nb <- stats::aggregate(year ~ station_number + period,
                       kept[kept$variable == "q_mean" & kept$year %in% baseline, ], length)
nb <- merge(expand.grid(station_number = stations, period = windows$window,
                        stringsAsFactors = FALSE), nb, all.x = TRUE)
nb$year[is.na(nb$year)] <- 0L
thin <- nb[nb$year < min_baseline, ]

# departure and trend from cd, one station per call
cols_flag <- c("variable", "period", "year", "frac_ice", "frac_provisional")
dep <- do.call(rbind, lapply(stations, function(st) {
  ok <- nb$period[nb$station_number == st & nb$year >= min_baseline]
  x <- kept[kept$station_number == st & kept$period %in% ok, ]
  x$station_number <- NULL
  a <- cd::cd_anomaly(x, suppressWarnings(cd::cd_baseline(x, baseline)))
  a <- merge(a, x[cols_flag], by = c("variable", "period", "year"))
  tr <- cd::cd_trend(a, trend_start = trend_start)
  list(anomaly = cbind(station_number = st, a), trend = cbind(station_number = st, tr))
}))
anom <- do.call(rbind, dep[, "anomaly"])
trend <- as.data.frame(do.call(rbind, dep[, "trend"]))
```

Fish use a river at set times of year: Chinook migrate up in early summer and
spawn in late summer, and their eggs incubate in the gravel through winter. A
low-flow year matters most when it falls inside one of those windows. This
vignette takes two hydrometric stations in the Bulkley watershed,
`r station_label[["08EE013"]]` and `r station_label[["08EE003"]]`, and asks how
flow in each species window compares with `r base_text`, and which way it is
trending.

## The recipe

These are the calls that produced the numbers. They need HYDAT, the
water-temp-bc archive and the ECCC real-time feed, so the vignette runs on a
cached copy of their output (see Cached inputs, below).

```{r recipe, echo = TRUE, eval = FALSE}
library(wet)

# HYDAT, then ECCC provisional, then real-time, per station
daily <- wet_station_daily(c("08EE013", "08EE003"))

windows <- data.frame(window = c("CH migration", "CH spawning"),
                      start = c("05-01", "08-01"), end = c("08-01", "09-15"))
stats <- wet_window_stats(daily, windows, stats = c("mean", "min7"))

# departure and trend are cd's, one station per call
buck <- stats[stats$station_number == "08EE013", names(stats) != "station_number"]
departure <- cd::cd_anomaly(buck, cd::cd_baseline(buck, 1981:2010))
trend <- cd::cd_trend(departure, trend_start = c(1981, 2000))
```

## What the record is made of

```{r record-data}
rec <- daily
rec$year <- as.integer(format(rec$date, "%Y"))
rec$class <- ifelse(rec$source == "hydat", ifelse(rec$symbol %in% "B", "hydat_ice", "hydat_open"),
                    rec$source)
rec <- stats::aggregate(list(days = rec$date), rec[c("station_number", "year", "class")], length)
rec$class <- factor(rec$class, names(pal_source$values))
buck_ice <- with(daily[daily$station_number == "08EE013" & daily$source == "hydat", ],
                 mean(symbol %in% "B"))
# a year is year-round when its winter is gauged: at least 80% of January and
# February has flow. Days in total would miss a current year, or a summer gap
w3 <- daily[daily$station_number == "08EE003" & format(daily$date, "%m") %in% c("01", "02"), ]
w3 <- table(as.integer(format(w3$date, "%Y")))
full <- as.integer(names(w3)[w3 >= 0.8 * 59])
cap_record <- sprintf(paste(
  "Days with flow per year, by source. HYDAT's ice-affected days (symbol B) are an estimate,",
  "not a measurement: %s of Buck Creek's approved days.",
  "Bulkley River near Houston is gauged through the winter only in %s."
), pct(100 * buck_ice), years_text(full))
```

Every number below rests on this record. Approved HYDAT flows end in
`r format(max(daily$date[daily$source == "hydat"]), "%B %Y")`; provisional and real-time flows carry it to
`r day_text(max(daily$date))`. Winters are mostly ice estimates, and
one station was a seasonal gauge for most of its history.

```{r record, fig.height = 4.5, fig.cap = cap_record}
ggplot(by_station(rec), aes(year, days, fill = class)) +
  geom_col(width = 1) +
  facet_wrap(~ station_number, ncol = 1, labeller = as_labeller(station_label)) +
  scale_fill_manual(values = pal_source$values, labels = pal_source$labels, name = NULL) +
  labs(x = NULL, y = "Days with flow")
```

## The hydrograph and the windows

```{r hydro-data}
doy_of <- function(date) {
  j <- as.integer(format(date, "%j"))
  y <- as.integer(format(date, "%Y"))
  leap <- (y %% 4 == 0 & y %% 100 != 0) | y %% 400 == 0
  ifelse(leap & j > 59, j - 1L, j)
}
h <- daily[format(daily$date, "%m-%d") != "02-29" & daily$q_m3s > 0, ]
h$doy <- doy_of(h$date)
h$year <- as.integer(format(h$date, "%Y"))

hb <- h[h$status == "approved" & h$year %in% baseline, ]
band <- do.call(rbind, lapply(split(hb, list(hb$station_number, hb$doy), drop = TRUE), function(d) {
  q <- stats::quantile(d$q_m3s, c(0.25, 0.5, 0.75), names = FALSE)
  data.frame(station_number = d$station_number[1], doy = d$doy[1], n = nrow(d),
             lo = q[1], med = q[2], hi = q[3])
}))
band <- band[band$n >= min_baseline, ]
lines <- h[h$year %in% recent, ]
lines$year <- factor(lines$year, names(pal_year$values))
lines$status <- factor(lines$status, c("approved", "provisional"))

# species windows as strips below the data, on the same log axis
floor_q <- vapply(stations, function(st) {
  min(band$lo[band$station_number == st], lines$q_m3s[lines$station_number == st])
}, 0)
strip <- do.call(rbind, lapply(stations, function(st) {
  do.call(rbind, lapply(seq_len(nrow(windows)), function(k) {
    s <- doy_of(as.Date(paste0("2001-", windows$start[k])))
    e <- doy_of(as.Date(paste0("2001-", sub("02-29", "02-28", windows$end[k]))))
    seg <- if (e >= s) data.frame(x = s, xend = e) else data.frame(x = c(s, 1), xend = c(365, e))
    cbind(station_number = st, window = windows$window[k], seg,
          y = floor_q[[st]] / 1.45^k)
  }))
}))
strip_lab <- strip[!duplicated(strip[c("station_number", "window")]), ]
month_start <- doy_of(as.Date(sprintf("2001-%02d-01", 1:12)))

spawn <- anom[anom$period == "CH spawning" & anom$variable == "q_mean" & anom$year %in% recent, ]
# the sentence below claims every recent year at both stations, all below normal
stopifnot(nrow(spawn) == length(recent) * length(stations), all(spawn$anomaly < 0))
cap_hydro <- sprintf(paste(
  "Daily flow, %d–%d (lines; dashed where provisional), over the %d–%d median and middle half",
  "(drawn only on days with at least %d years). Species windows run beneath. Log scale."
), min(recent), max(recent), min(baseline), max(baseline), min_baseline)
```

This is the key figure. Each recent year is drawn over the `r base_text` normal
for that day, with the species windows underneath. In the Chinook spawning
window, both stations ran `r pct(max(spawn$anomaly))`–`r pct(min(spawn$anomaly))`
below normal in every year from `r min(recent)` to `r max(recent)`. The
dashed winter lines are provisional readings under ice; they are
the reason those years are dropped below.

```{r hydrograph, fig.height = 8, fig.cap = cap_hydro}
band <- by_station(band)
lines <- by_station(lines)
strip <- by_station(strip)
strip_lab <- by_station(strip_lab)
ggplot() +
  geom_ribbon(data = band, aes(doy, ymin = lo, ymax = hi, fill = "band")) +
  geom_line(data = band, aes(doy, med, colour = "median"), linewidth = 0.5) +
  geom_line(data = lines, aes(doy, q_m3s, colour = year, linetype = status,
                              group = interaction(year, status)), linewidth = 0.45) +
  geom_segment(data = strip, aes(x = x, xend = xend, y = y, yend = y),
               colour = pal_hydro$values[["window"]], linewidth = 1.6) +
  geom_text(data = strip_lab, aes(x = -4, y = y, label = window), hjust = 1, size = 2.6) +
  facet_wrap(~ station_number, ncol = 1, scales = "free_y",
             labeller = as_labeller(station_label)) +
  # no ticks among the strips: they are rows, not flows
  scale_y_log10(breaks = function(l) {
    b <- 10^(floor(log10(l[1])):ceiling(log10(l[2])))
    b[b >= l[1] * 1.45^(nrow(windows) + 0.5)]
  }, labels = function(x) format(x, drop0trailing = TRUE, scientific = FALSE)) +
  scale_x_continuous(breaks = month_start, labels = month.abb, limits = c(-80, 366),
                     expand = c(0, 0)) +
  scale_fill_manual(values = pal_hydro$values["band"], labels = pal_hydro$labels["band"],
                    name = NULL) +
  scale_colour_manual(values = c(pal_hydro$values["median"], pal_year$values),
                      labels = c(pal_hydro$labels["median"], pal_year$labels), name = NULL) +
  scale_linetype_manual(values = c(approved = "solid", provisional = "22"),
                        labels = c(approved = "Approved", provisional = "Provisional"), name = NULL) +
  labs(x = NULL, y = "Discharge (m³/s)") +
  guides(fill = guide_legend(order = 1), colour = guide_legend(order = 2, nrow = 1),
         linetype = guide_legend(order = 3)) +
  theme(panel.grid.major.x = element_blank(), legend.box = "vertical",
        legend.spacing.y = unit(0, "pt"), legend.margin = margin(0, 0, 0, 0))
```

## Departure, window by window

```{r heat-data}
heat <- merge(stats[stats$variable == "q_mean" & stats$year >= min(baseline),
                    c("station_number", "period", "year", "frac_ice", "frac_provisional", "dropped")],
              anom[anom$variable == "q_mean", c("station_number", "period", "year", "anomaly")],
              all.x = TRUE)
heat$class <- as.character(cut(heat$anomaly, c(-Inf, -75, -50, -25, 0, 25, 50, 100, Inf),
                               labels = paste0("d", 1:8)))
heat$class[heat$dropped] <- "dropped"
heat$class[paste(heat$station_number, heat$period) %in%
             paste(thin$station_number, thin$period)] <- "no_baseline"
heat$class <- factor(heat$class, names(pal_dep$values))
heat$period <- factor(heat$period, rev(windows$window))
heat$flag <- ifelse(heat$frac_provisional > 0 & !heat$dropped, "provisional", NA)
# ice is a property of a window, not of a year: only the incubation windows
# carry much, so it is stated once rather than marked per cell
ice_win <- stats::aggregate(frac_ice ~ period, heat[heat$station_number == "08EE013", ], stats::median)
ice_win <- ice_win[ice_win$frac_ice > 0.25, ]
stopifnot(nrow(ice_win) > 0)
ice_text <- paste(sprintf("%s of the %s days", pct(100 * ice_win$frac_ice), ice_win$period),
                  collapse = " and ")

# from the strip's own classes, so the sentence and the figure cannot disagree
# (a window with too few baseline years shows as that, whatever else applies)
drop_text <- vapply(stations, function(st) {
  d <- heat[heat$station_number == st & heat$class %in% "dropped", ]
  if (!nrow(d)) return("none")
  w <- intersect(windows$window, as.character(d$period))
  paste(w, vapply(w, function(x) years_text(d$year[d$period == x]), ""), collapse = "; ")
}, "")
drop_sentence <- if (length(unique(drop_text)) == 1) {
  paste("the same at both stations:", drop_text[1])
} else {
  paste(sprintf("at %s, %s", station_label, drop_text), collapse = "; and ")
}
pw <- daily[daily$station_number == "08EE013" & format(daily$date, "%m") %in% c("12", "01", "02"), ]
pw$month <- format(pw$date, "%Y-%m")
pm <- stats::aggregate(q_m3s ~ month + status, pw, mean)
prov_range <- range(pm$q_m3s[pm$status == "provisional"])
appr_max <- max(pm$q_m3s[pm$status == "approved" & substr(pm$month, 1, 4) >= min(baseline)])
cap_heat <- sprintf(paste(
  "Window mean flow in %% of the %d–%d normal. A triangle marks a window-year resting on",
  "provisional days. Blank cells have under %s of their days, or none."
), min(baseline), max(baseline), pct(100 * formals(wet_window_stats)$min_frac))
```

Each cell is one window in one year, as a percentage of its `r base_text` mean.
Two kinds of cell carry no departure. A window at a station with fewer than
`r min_baseline` baseline years has no normal to compare against. And a window
touching November–April drops every year with a provisional day: Buck Creek's
provisional December–February means were
`r sprintf("%.1f–%.1f", prov_range[1], prov_range[2])` m³/s, where no approved
winter month since `r min(baseline)` exceeded `r sprintf("%.1f", appr_max)` m³/s. The
window-years dropped for that reason are `r drop_sentence`. Winter windows
also rest on ice estimates: in Buck Creek's median year, `r ice_text` are
ice-affected.

```{r heat, fig.height = 6.5, fig.cap = cap_heat}
heat <- by_station(heat)
ggplot(heat, aes(year, period)) +
  geom_tile(aes(fill = class), colour = "white", linewidth = 0.2) +
  geom_point(data = heat[!is.na(heat$flag), ], aes(shape = flag), size = 0.7) +
  facet_wrap(~ station_number, ncol = 1, labeller = as_labeller(station_label)) +
  scale_fill_manual(values = pal_dep$values, labels = pal_dep$labels, name = NULL, drop = FALSE) +
  scale_shape_manual(values = c(provisional = 2), labels = c(provisional = "Provisional"),
                     name = NULL) +
  scale_x_continuous(expand = c(0, 0)) +
  labs(x = NULL, y = NULL) +
  guides(fill = guide_legend(ncol = 4, order = 1), shape = guide_legend(order = 2)) +
  theme(panel.grid = element_blank(), legend.box = "vertical")
```

## Trends

```{r trend-data}
tr <- trend[trend$variable %in% c("q_mean", "q_min7"), ]
tr$variable <- factor(tr$variable, c("q_mean", "q_min7"),
                      c("Window mean", "Lowest 7-day mean"))
tr$period <- factor(tr$period, rev(windows$window))
tr$trend_start <- factor(tr$trend_start, names(pal_trend$values))
tr$sig <- factor(ifelse(tr$mk_pvalue < alpha, "sig", "ns"), c("sig", "ns"))
# windows with the same dates give the same slope, so count each result once
tr$dates <- paste(windows$start, windows$end)[match(tr$period, windows$window)]
key <- paste(windows$start, windows$end)
groups <- split(windows$window, key)[unique(key[duplicated(key)])]
same_text <- if (length(groups)) {
  paste0(paste(vapply(groups, paste, "", collapse = " and "), collapse = "; "),
         " share their dates, so each set counts once. ")
} else {
  ""
}
distinct <- tr[!duplicated(tr[c("station_number", "variable", "trend_start", "dates")]), ]
n_tested <- nrow(distinct)
n_sig <- sum(distinct$sig == "sig")
n_neg <- sum(distinct$sig == "sig" & distinct$slope < 0)
sign_text <- if (n_sig == 0) {
  ""
} else if (n_neg == n_sig) {
  ", all of them negative"
} else {
  sprintf(", %d of them negative", n_neg)
}
# where 08EE003's "from 2000" trends really start: the first year of each fitted series
fit <- trend[trend$station_number == "08EE003" & trend$trend_start == trend_start[2] &
               trend$variable %in% c("q_mean", "q_min7"), ]
a3 <- anom[anom$station_number == "08EE003" & anom$year >= trend_start[2] & !is.na(anom$anomaly), ]
a3 <- a3[paste(a3$period, a3$variable) %in% paste(fit$period, fit$variable), ]
start_003 <- tapply(a3$year, paste(a3$period, a3$variable), min)
stopifnot(length(start_003) == nrow(fit))
since_2000_003 <- if (length(unique(start_003)) == 1) {
  as.character(start_003[1])
} else {
  sub(", ([^,]*)$", " or \\1", paste(sort(unique(start_003)), collapse = ", "))
}
cap_trend <- sprintf(paste(
  "Theil-Sen slope of the departure, in percentage points of normal per year, from %s.",
  "Filled where the Mann-Kendall p is below %s. A window with no point has under %d baseline years."
), paste(trend_start, collapse = " and from "), alpha, min_baseline)
```

The slope of each window's departure, `r trend_text`, for the window
mean and its lowest seven-day mean. `r same_text`Of `r n_tested` distinct
slopes, `r n_sig` are significant at p < `r alpha`, where chance alone would give
about `r round(alpha * n_tested)``r sign_text`. At Bulkley River near Houston,
"from `r trend_start[2]`" starts in `r since_2000_003`, depending on the window,
because it has no window-years from `r trend_start[2]` until then.

```{r trend, fig.height = 6.5, fig.cap = cap_trend}
tr <- by_station(tr)
ggplot(tr, aes(slope, period, colour = trend_start, shape = sig)) +
  geom_vline(xintercept = 0, colour = pal_hydro$values[["median"]], linewidth = 0.3) +
  geom_point(position = position_dodge(width = 0.6), size = 1.8) +
  facet_grid(station_number ~ variable, labeller = labeller(station_number = station_label)) +
  scale_colour_manual(values = pal_trend$values, labels = pal_trend$labels, name = NULL) +
  scale_shape_manual(values = c(sig = 16, ns = 1),
                     labels = c(sig = paste("p <", alpha), ns = paste("p ≥", alpha)), name = NULL) +
  labs(x = "Slope (percentage points of normal per year)", y = NULL)
```

## What it does not show

```{r limits}
n_unsourced <- sum(is.na(lh$source))
b3 <- nb$year[nb$station_number == "08EE003" & nb$period == "CH incubation"]
thin_003 <- sub(", ([^,]*)$", " and \\1",
                paste(intersect(windows$window, thin$period[thin$station_number == "08EE003"]),
                      collapse = ", "))
stopifnot(nzchar(thin_003))
sp <- kept[kept$station_number == "08EE013" & kept$period == "CH spawning" &
             kept$variable == "q_mean" & kept$year %in% baseline, "value"]
```

- **Provisional winters are ice readings.** ECCC's provisional archive
  records no ice symbol, so nothing marks those days. The drop above is a
  rule, not a correction.
- **Bulkley River near Houston was a seasonal gauge for most of its record**,
  gauged through the winter only in `r years_text(full)`. Its `r thin_003` windows have under
  `r min_baseline` baseline years (Chinook incubation has `r b3`), so no
  departure.
- **A mean-based normal is skewed by wet years.** At Buck Creek the
  `r base_text` Chinook spawning mean is `r sprintf("%.2f", mean(sp))` m³/s, and the median
  year `r sprintf("%.2f", stats::median(sp))` m³/s, so a typical year already reads
  `r pct(100 * (stats::median(sp) / mean(sp) - 1))` below normal.
- **The windows come from the literature, not from observation.** They are
  Bulkley timings from NGE's life-history table; `r n_unsourced` of
  `r nrow(lh)` have no source yet and await confirmation with knowledge
  holders.

**Cached inputs.** HYDAT release `r prov$hydat`, with provisional and
real-time flows retrieved `r day_text(prov$retrieved)`; life-history
table at commit `r substr(prov$knowledge_sha, 1, 7)`. Regenerated by
`data-raw/station_vignette_data.R`.

```{r wordcount, include = FALSE}
f <- knitr::current_input()
if (!is.null(f)) {
  x <- readLines(f, warn = FALSE)
  y <- utils::tail(x, -utils::head(which(x == "---"), 2)[2])   # drop the YAML header
  fence <- cumsum(grepl("^```", y)) %% 2 == 1
  y <- y[!fence & !grepl("^```", y)]
  y <- gsub("`r [^`]*`", "", y)          # inline code is not body prose
  y <- y[!grepl("^\\s*$", y)]
  n <- sum(lengths(strsplit(trimws(y), "\\s+")))
  message("body prose: ", n, " words")
  cap <- 600   # 485 at #27
  if (n > cap) stop("body prose is ", n, " words; the cap is ", cap)
}
```
