---
title: "Chinook Flow at Two Stations"
output: bookdown::html_vignette2
vignette: >
  %\VignetteIndexEntry{Chinook Flow at Two Stations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.width = 7,
  out.width = "100%",
  dpi = 150
)
```

```{r load}
library(wet)
library(ggplot2)
sf::sf_use_s2(FALSE)

vd <- function(f) system.file("vignette-data", f, package = "wet")
daily <- readRDS(vd("station_daily.rds"))
prov <- attr(daily, "provenance")
map <- readRDS(vd("station_map.rds"))
lh <- utils::read.csv(vd("life_history_bulk.csv"), stringsAsFactors = FALSE, na.strings = "")

to_title <- function(x) {
  vapply(strsplit(tolower(x), " "), function(w) {
    big <- !w %in% c("at", "the", "near", "of")
    w[big] <- paste0(toupper(substring(w[big], 1, 1)), substring(w[big], 2))
    paste(w, collapse = " ")
  }, "")
}
station_label <- stats::setNames(
  sprintf("%s (%s)", to_title(prov$stations$station_name), prov$stations$station_number),
  prov$stations$station_number
)
stations <- names(station_label)

# Chinook windows that stay in open water: a window touching November-April
# rests on flows read under ice, so it is left out rather than corrected
ice_months <- c(11:12, 1:4)
touches_ice <- function(start, end) {
  a <- as.Date(paste0("2001-", start))
  b <- as.Date(paste0(2001 + (end < start), "-", sub("02-29", "02-28", end)))
  any(as.integer(format(seq(a, b, by = "day"), "%m")) %in% ice_months)
}
ch <- lh[lh$species_code == "CH", ]
open_water <- !mapply(touches_ice, ch$start, ch$end)
stopifnot(identical(ch$life_stage[open_water], c("migration", "spawning", "fry_migration")))
out <- ch[!open_water, ]
ch <- ch[open_water, ]
stage <- sub("_", " ", ch$life_stage)
windows <- data.frame(window = paste0(toupper(substring(stage, 1, 1)), substring(stage, 2)),
                      start = ch$start, end = ch$end)
md_text <- function(md) {
  d <- as.Date(paste0("2001-", md))
  paste(format(d, "%b"), as.integer(format(d, "%d")))
}
window_label <- stats::setNames(sprintf("%s\n%s – %s", windows$window, md_text(windows$start),
                                        md_text(windows$end)), windows$window)
span_text <- function(start, end) paste0(md_text(start), " – ", md_text(end))
days_of <- function(k) {
  seq(as.Date(paste0("2001-", windows$start[k])), as.Date(paste0("2001-", windows$end[k])), by = "day")
}
win_days <- stats::setNames(vapply(seq_len(nrow(windows)), function(k) length(days_of(k)), 0L),
                            windows$window)
# a window-year counts when this share of its days has flow (wet_window_stats())
need_text <- paste0(round(100 * formals(wet_window_stats)$min_frac), "%")
left_out <- sprintf("%s (%s)", sub("_", " ", out$life_stage), span_text(out$start, out$end))
w_start <- stats::setNames(md_text(windows$start), windows$window)
# the recipe chunk below types these dates; fail here if the table moves under it
stopifnot(identical(windows$start, c("05-01", "08-01", "07-15")),
          identical(windows$end, c("08-01", "09-15", "09-07")))

baseline <- 1981:2010
min_baseline <- 10
trend_start <- c(1981, 2000)
alpha <- 0.05
base_text <- paste0(min(baseline), "–", max(baseline))
trend_text <- paste("from", trend_start, collapse = " and ")
last_year <- as.integer(format(max(daily$date), "%Y"))
recent <- (last_year - 4):last_year
recent_text <- paste0(min(recent), "–", max(recent))
# every window must be over in the last year, or its last bar is a part-year
stopifnot(all(as.Date(paste0(last_year, "-", windows$end)) <= max(daily$date)))

# colours come from the gq registries, so no hex literal appears in this
# vignette; assert each class set, because a malformed CSV row adds a layer silently
reg <- gq::gq_reg_custom(system.file("cartography", "wet_station.csv", package = "wet"))
reg_main <- gq::gq_reg_main()
classes <- function(layer, expect) {
  x <- gq::gq_tmap_classes(reg$layers[[layer]])
  stopifnot(identical(names(x$values), expect))
  list(values = x$values, labels = stats::setNames(x$labels, names(x$values)))
}
pal_hydro <- classes("hydrograph", c("band", "median", "window"))
pal_hydro$labels <- sub("^Baseline", base_text, pal_hydro$labels)
# ordinal classes, so a data refresh needs no registry edit
pal_year <- classes("recent_year", c("latest_4", "latest_3", "latest_2", "latest_1", "latest"))
names(pal_year$values) <- names(pal_year$labels) <- pal_year$labels <- as.character(recent)
pal_dep <- classes("departure_class", paste0("d", 1:8))
pal_trend <- classes("trend_start", as.character(trend_start))
pal_side <- classes("recent_departure", c("below", "above"))
pal_map <- classes("station_map", c("catchment_main", "catchment_buck", "stream_main", "station"))
# labels that restate a variable are built from it, not read from the registry
pal_trend$labels[] <- paste("Since", names(pal_trend$values))
pal_side$labels[] <- paste(c("Below", "Above"), "the", base_text, "mean")

theme_set(theme_minimal(base_size = 10) +
            theme(legend.position = "bottom", panel.grid.minor = element_blank()))
pct <- function(x) paste0(round(abs(x)), "%")
day_text <- function(d) paste(as.integer(format(d, "%d")), format(d, "%B %Y"))
by_station <- function(d) {
  d$station_number <- factor(d$station_number, stations)
  d
}
by_window <- function(d, rev = FALSE) {
  lv <- if (rev) rev(windows$window) else windows$window
  d$period <- factor(d$period, lv)
  d
}
years_text <- function(y) {
  y <- sort(unique(y))
  runs <- split(y, cumsum(c(1, diff(y) != 1)))
  one <- function(r) if (length(r) > 1) paste0(r[1], "–", r[length(r)]) else as.character(r)
  sub(", ([^,]*)$", " and \\1", paste(vapply(runs, one, ""), collapse = ", "))
}
and_text <- function(x) sub(", ([^,]*)$", " and \\1", paste(x, collapse = ", "))
```

```{r stats}
stats <- wet_window_stats(daily, windows, stats = c("mean", "min7"))

# baseline years per station and window; every one must clear min_baseline,
# because nothing below says what happens to a window that does not
qb <- stats[stats$variable == "q_mean" & stats$year %in% baseline, ]
nb <- merge(expand.grid(station_number = stations, period = windows$window,
                        stringsAsFactors = FALSE),
            stats::aggregate(year ~ station_number + period, qb, length), all.x = TRUE)
nb$year[is.na(nb$year)] <- 0L
stopifnot(all(nb$year >= min_baseline))
# 08EE003's baseline years, the same in every window (the prose states them once)
y003 <- split(qb$year[qb$station_number == "08EE003"], qb$period[qb$station_number == "08EE003"])
stopifnot(length(unique(lapply(y003, sort))) == 1)
base_003 <- years_text(y003[[1]])
miss_003 <- setdiff(baseline, y003[[1]])
# the caption and the limits say 08EE003's baseline is shorter than the period
stopifnot(length(miss_003) > 0)
# baseline years missing at every station and window, for the limits bullet
miss_text <- vapply(stations, function(st) {
  m <- vapply(windows$window, function(w) {
    paste(setdiff(baseline, qb$year[qb$station_number == st & qb$period == w]), collapse = " ")
  }, "")
  m <- m[nzchar(m)]
  if (!length(m)) return(NA_character_)
  parts <- vapply(unique(m), function(k) {
    w <- names(m)[m == k]
    where <- if (length(w) == nrow(windows)) "every window" else tolower(and_text(w))
    sprintf("%s in %s", years_text(as.integer(strsplit(k, " ")[[1]])), where)
  }, "")
  sprintf("%s lacks %s", sub(" \\(.*", "", station_label[[st]]), paste(parts, collapse = ", and "))
}, "")
miss_text <- paste(miss_text[!is.na(miss_text)], collapse = "; ")

# baseline, departure and trend from cd, one station per call
cols_flag <- c("variable", "period", "year", "value", "n_days", "frac_provisional")
dep <- lapply(stations, function(st) {
  x <- stats[stats$station_number == st, ]
  x$station_number <- NULL
  # cd warns of baseline years with no data; base_003 states them instead
  b <- suppressWarnings(cd::cd_baseline(x, baseline))
  a <- merge(cd::cd_anomaly(x, b), x[cols_flag], by = c("variable", "period", "year"))
  list(baseline = cbind(station_number = st, b),
       anomaly = cbind(station_number = st, a),
       trend = cbind(station_number = st, cd::cd_trend(a, trend_start = trend_start)))
})
pick <- function(k) do.call(rbind, lapply(dep, `[[`, k))
base <- pick("baseline")
anom <- pick("anomaly")
trend <- as.data.frame(pick("trend"))
```

Chinook use the Bulkley River in summer. Adults migrate upstream from
`r w_start[["Migration"]]` and spawn from `r w_start[["Spawning"]]`, and fry move
downstream from `r w_start[["Fry migration"]]`. This vignette asks how flow in
those three windows over the last five years, `r recent_text`, compares with
`r base_text` at two hydrometric stations near Houston. Chinook
`r and_text(left_out)` reach into November–April, when the gauges can read
under ice, so they are left out.

## Where the stations are

```{r map-data}
catch <- map$catchments
catch$part <- ifelse(catch$station_number == "08EE013", "catchment_buck", "catchment_main")
catch <- catch[order(catch$part != "catchment_main"), ]   # the larger one first, beneath
buck_share <- with(sf::st_drop_geometry(catch),
                   area_km2[station_number == "08EE013"] / area_km2[station_number == "08EE003"])
# the map claims Buck Creek's catchment lies inside the Bulkley gauge's
inside <- sf::st_area(sf::st_intersection(sf::st_geometry(catch[catch$part == "catchment_buck", ]),
                                          sf::st_geometry(catch[catch$part == "catchment_main", ])))
stopifnot(as.numeric(inside) / as.numeric(sf::st_area(catch[catch$part == "catchment_buck", ])) > 0.99)
# the prose puts both stations "near Houston", and Buck Creek's mouth "in the town"
d_town <- stats::setNames(as.numeric(sf::st_distance(map$stations, map$places)), map$stations$station_number)
stopifnot(all(d_town < 10000), d_town[["08EE013"]] < 1000)

map_w <- 7
map_h <- 6.5
bb <- sf::st_bbox(sf::st_buffer(sf::st_union(catch), 3000))
# room below the catchments for the legend and scale bar, so neither covers them
bb[["ymin"]] <- bb[["ymin"]] - 0.22 * (bb[["ymax"]] - bb[["ymin"]])
bb <- gq::gq_bbox_aspect(bb, asp = map_w / map_h)
lim_x <- c(bb[["xmin"]], bb[["xmax"]])
lim_y <- c(bb[["ymin"]], bb[["ymax"]])
dx <- diff(lim_x)
dy <- diff(lim_y)

st <- map$stations
xy <- sf::st_coordinates(st)
# 08EE003 sits just west of Buck Creek's mouth and near the frame's west edge:
# its label goes above and right, Buck Creek's to the right, Houston's below left
west <- xy[, 1] == min(xy[, 1])
st_lab <- data.frame(x = xy[, 1] + ifelse(west, -0.01, 0.015) * dx,
                     y = xy[, 2] + ifelse(west, 0.03, 0) * dy,
                     label = station_label[st$station_number], vjust = ifelse(west, 0, 0.5))
pl <- sf::st_coordinates(map$places)
# the town point sits under Buck Creek's marker, so the town is a label below it
pl_lab <- data.frame(x = pl[, 1], y = pl[, 2] - 0.035 * dy, label = map$places$name)

stream_sty <- reg_main$layers$streams_all$classification$classes$WA24111110
lake_sty <- reg_main$layers$lake
town_sty <- reg_main$layers$town
main_streams <- map$streams[map$streams$main, ]

# scale bar in the lower right, its length from gq
sb_km <- max(gq::gq_scale_breaks(bb))
sb <- data.frame(x = lim_x[2] - 0.05 * dx - sb_km * 1000, y = lim_y[1] + 0.05 * dy)
sb$xend <- sb$x + sb_km * 1000

# keymap: BC with the map extent, in the upper right
key <- ggplot() +
  geom_sf(data = map$bc, fill = "white", colour = pal_hydro$values[["median"]], linewidth = 0.2) +
  geom_sf(data = sf::st_as_sfc(bb), fill = NA, colour = pal_map$values[["station"]], linewidth = 0.6) +
  theme_void() +
  theme(panel.background = element_rect(fill = "white", colour = pal_hydro$values[["median"]],
                                        linewidth = 0.3))
key_w <- 0.24 * dx
key_h <- key_w * 1.05

cap_map <- sprintf(paste(
  "The two stations and the land each one drains. Buck Creek's catchment (%s km²) lies inside",
  "the Bulkley gauge's (%s km²), so the two records are not independent."
), format(round(catch$area_km2[catch$part == "catchment_buck"]), big.mark = ","),
format(round(catch$area_km2[catch$part == "catchment_main"]), big.mark = ","))
```

Both stations are near Houston. Buck Creek joins the Bulkley in the town, and the
Bulkley gauge sits downstream, so it measures Buck Creek's flow as well. Buck Creek
drains `r pct(100 * buck_share)` of the area above the Bulkley gauge.

```{r map, fig.width = map_w, fig.height = map_h, fig.cap = cap_map}
ggplot() +
  geom_sf(data = catch, aes(fill = part), colour = pal_hydro$values[["median"]], linewidth = 0.25,
          key_glyph = "rect") +
  geom_sf(data = map$lakes, fill = lake_sty$fill$color, colour = lake_sty$stroke$color,
          linewidth = 0.15) +
  geom_sf(data = map$streams[!map$streams$main, ], colour = stream_sty$color, linewidth = 0.35) +
  geom_sf(data = main_streams, aes(colour = "stream_main"), linewidth = 0.9, key_glyph = "path") +
  geom_label(data = pl_lab, aes(x, y, label = label), colour = town_sty$label$color,
             fontface = "bold", size = 3.6, linewidth = 0, fill = "white", alpha = 0.85,
             label.padding = unit(1.5, "pt")) +
  geom_sf(data = st, aes(shape = "station"), fill = pal_map$values[["station"]], colour = "black",
          size = 3, stroke = 0.4) +
  geom_label(data = st_lab, aes(x, y, label = label, vjust = vjust), hjust = 0, size = 3.4,
             linewidth = 0, fill = "white", alpha = 0.85, label.padding = unit(1.5, "pt")) +
  geom_segment(data = sb, aes(x = x, xend = xend, y = y, yend = y), linewidth = 1) +
  geom_text(data = sb, aes(x = (x + xend) / 2, y = y, label = paste(sb_km, "km")),
            vjust = -0.7, size = 3.4) +
  annotation_custom(ggplotGrob(key), xmin = lim_x[2] - key_w - 0.02 * dx, xmax = lim_x[2] - 0.02 * dx,
                    ymin = lim_y[2] - key_h - 0.02 * dy, ymax = lim_y[2] - 0.02 * dy) +
  scale_fill_manual(values = pal_map$values[c("catchment_main", "catchment_buck")],
                    labels = pal_map$labels[c("catchment_main", "catchment_buck")], name = NULL) +
  scale_colour_manual(values = pal_map$values["stream_main"], labels = pal_map$labels["stream_main"],
                      name = NULL) +
  scale_shape_manual(values = c(station = 21), labels = pal_map$labels["station"], name = NULL) +
  guides(shape = guide_legend(order = 1, override.aes = list(fill = pal_map$values[["station"]])),
         colour = guide_legend(order = 2), fill = guide_legend(order = 3)) +
  coord_sf(xlim = lim_x, ylim = lim_y, expand = FALSE, datum = NA) +
  theme_void(base_size = 12) +
  theme(legend.position = "inside", legend.position.inside = c(0.02, 0.02),
        legend.justification = c(0, 0), legend.box.just = "left", legend.box = "vertical",
        legend.box.background = element_rect(fill = "white", colour = NA),
        legend.background = element_rect(fill = "white", colour = NA),
        legend.margin = margin(4, 6, 4, 6), legend.spacing.y = unit(0, "pt"),
        panel.border = element_rect(fill = NA, colour = pal_hydro$values[["median"]]))
```

## The recipe

These are the calls that produced the numbers. They need HYDAT, the
water-temp-bc archive and the ECCC real-time feed, so the vignette runs on a
cached copy of their output (see Cached inputs, below).

```{r recipe, echo = TRUE, eval = FALSE}
library(wet)

# HYDAT, then ECCC provisional, then real-time, per station
daily <- wet_station_daily(c("08EE013", "08EE003"))

windows <- data.frame(window = c("Migration", "Spawning", "Fry migration"),
                      start = c("05-01", "08-01", "07-15"), end = c("08-01", "09-15", "09-07"))
stats <- wet_window_stats(daily, windows, stats = c("mean", "min7"))

# baseline, departure and trend are cd's, one station per call
buck <- stats[stats$station_number == "08EE013", names(stats) != "station_number"]
base <- cd::cd_baseline(buck, 1981:2010)
departure <- cd::cd_anomaly(buck, base)
trend <- cd::cd_trend(departure, trend_start = c(1981, 2000))
```

## The last five years against `r base_text`

```{r recent-data}
kf <- anom[anom$variable == "q_mean" & anom$year %in% recent, ]
stopifnot(nrow(kf) == length(recent) * nrow(windows) * length(stations), !anyNA(kf$anomaly))
kf$pct <- 100 + kf$anomaly
kf$side <- ifelse(kf$pct < 100, "below", "above")
kf$status <- ifelse(kf$frac_provisional > 0, "provisional", "approved")

# the median baseline year, as % of cd's baseline mean: in a skewed record a
# typical year sits under the mean, and the figure has to show that
bq <- base[base$variable == "q_mean", ]
ref <- do.call(rbind, lapply(seq_len(nrow(bq)), function(i) {
  v <- qb$value[qb$station_number == bq$station_number[i] & qb$period == bq$period[i]]
  stopifnot(isTRUE(all.equal(mean(v), bq$baseline_mean[i])))
  data.frame(station_number = bq$station_number[i], period = bq$period[i],
             mean_q = bq$baseline_mean[i], median_pct = 100 * stats::median(v) / mean(v),
             q10 = stats::quantile(v, 0.1, names = FALSE), q_min = min(v))
}))
ref$label <- sprintf("Mean %s m³/s", vapply(signif(ref$mean_q, 2), format, "", drop0trailing = TRUE))

# the claims below, each computed from kf and ref
kr <- merge(kf, ref, by = c("station_number", "period"))
low_years <- recent[vapply(recent, function(y) all(kr$pct[kr$year == y] < 100), TRUE)]
stopifnot(length(low_years) >= 2)
low <- kr[kr$year %in% low_years, ]
# value is the window mean itself, carried from wet_window_stats() into anom
stopifnot(is.numeric(low[["value"]]), !anyNA(low[["value"]]))
n_low <- nrow(low)
n_dec <- sum(low$value < low$q10)
# the sentence says "but N of them": it needs at least one
stopifnot(n_dec > 0)
n_min <- sum(low$value < low$q_min)
worst <- kr[which.min(kr$pct), ]
up <- kr[kr$pct >= 100, ]
up_years <- sort(unique(up$year))
# the sentence names one wet year and the windows above normal at both stations
stopifnot(length(up_years) == 1)
up_both <- windows$window[vapply(windows$window, function(w) {
  all(stations %in% up$station_number[up$period == w])
}, TRUE)]
stopifnot(length(up_both) > 0)
agree <- stats::aggregate(side ~ period + year, kr, function(s) length(unique(s)) == 1)
med_range <- range(ref$median_pct)
# bars resting on fewer than all of their window's days
part <- kf[kf$n_days < win_days[kf$period], ]
stopifnot(nrow(part) > 0)   # the limits bullet says some bars have days missing
part_text <- paste(vapply(split(part, paste(part$station_number, part$year)), function(d) {
  sprintf("%s in %d (%s)", sub(" \\(.*", "", station_label[[d$station_number[1]]]), d$year[1],
          tolower(and_text(intersect(windows$window, d$period))))
}, ""), collapse = "; ")
part_share <- range(100 * part$n_days / win_days[part$period])
# "a typical year falling under a mean that wet years pull up": every median sits below the mean
stopifnot(round(med_range[2]) < 100)   # as printed

cap_recent <- sprintf(paste(
  "Mean flow in each Chinook window, %s, as a percentage of its %s mean (solid line, given in",
  "m³/s in each panel). The dashed line is the median %s year.",
  "At %s, the %s baseline rests on %s: no window in %s has %s of its days."
), recent_text, base_text, base_text, station_label[["08EE003"]], base_text, base_003,
years_text(miss_003), need_text)
```

Each bar is one window in one year. Its height is the window's mean flow as a
percentage of that window's `r base_text` mean.

Every window at both stations was below normal in `r years_text(low_years)`.
That is not simply a typical year falling under a mean that wet years pull up. A
median `r base_text` year sits at `r pct(med_range[1])`–`r pct(med_range[2])` of the
mean, but `r n_dec` of the `r n_low` bars from those years were drier than nine
`r base_text` years in ten, and `r n_min` were drier than any. The lowest was
`r tolower(worst$period)` at `r sub(" \\(.*", "", station_label[[worst$station_number]])` in
`r worst$year`, at `r pct(worst$pct)` of normal. `r up_years` was the exception,
with `r tolower(and_text(up_both))` above normal at both stations. For each window
and year, the two stations' bars fall on the same side of normal in
`r sum(agree$side)` of `r nrow(agree)` cases. That is partly by construction,
because one catchment holds the other.

```{r recent, fig.height = 5, fig.cap = cap_recent}
kf <- by_window(by_station(kf))
ref <- by_window(by_station(ref))
ymax <- max(kf$pct) * 1.2
refs <- rbind(data.frame(ref[c("station_number", "period")], y = 100, line = "mean"),
              data.frame(ref[c("station_number", "period")], y = ref$median_pct, line = "median"))
ggplot(kf, aes(year, pct)) +
  geom_col(aes(fill = side, alpha = status), width = 0.75) +
  geom_hline(data = refs, aes(yintercept = y, linetype = line), colour = pal_hydro$values[["median"]],
             linewidth = 0.4) +
  geom_text(aes(label = pct(pct)), vjust = -0.35, size = 2.7) +
  # the mean in m3/s, top right, clear of the 2022 bars
  geom_text(data = ref, aes(x = max(recent) + 0.45, y = ymax, label = label), hjust = 1,
            vjust = 1, size = 2.7, colour = pal_hydro$values[["median"]]) +
  facet_grid(station_number ~ period, labeller = labeller(
    station_number = as_labeller(sub(" \\(", "\n(", station_label)),
    period = as_labeller(window_label))) +
  scale_fill_manual(values = pal_side$values, labels = pal_side$labels, name = NULL) +
  scale_alpha_manual(values = c(approved = 1, provisional = 0.45),
                     labels = c(approved = "Approved", provisional = "Provisional"), name = NULL) +
  scale_linetype_manual(values = c(mean = "solid", median = "22"),
                        labels = c(mean = paste(base_text, "mean"),
                                   median = paste("Median", base_text, "year")), name = NULL) +
  scale_y_continuous(labels = function(x) paste0(x, "%"), limits = c(0, ymax),
                     expand = expansion(mult = c(0, 0))) +
  scale_x_continuous(breaks = recent) +
  labs(x = NULL, y = paste("Percent of", base_text, "mean")) +
  guides(fill = guide_legend(order = 1), alpha = guide_legend(
    order = 2, override.aes = list(fill = pal_side$values[["below"]])),
    linetype = guide_legend(order = 3, override.aes = list(linewidth = 0.5))) +
  theme(panel.grid.major.x = element_blank(), strip.text.y = element_text(angle = 0, hjust = 0),
        legend.box = "vertical", legend.spacing.y = unit(0, "pt"), legend.margin = margin(0, 0, 0, 0))
```

## The daily record

```{r hydro-data}
doy_of <- function(date) {
  j <- as.integer(format(date, "%j"))
  y <- as.integer(format(date, "%Y"))
  leap <- (y %% 4 == 0 & y %% 100 != 0) | y %% 400 == 0
  ifelse(leap & j > 59, j - 1L, j)
}
# May to October: the windows open on 1 May, and earlier days can still be ice readings
season <- doy_of(as.Date(c("2001-05-01", "2001-10-31")))
# a log axis drops zero flows; none fall in the season drawn, so a break means no record
zero <- daily$q_m3s == 0 & as.integer(format(daily$date, "%m")) %in% 5:10
stopifnot(!any(zero))
h <- daily[format(daily$date, "%m-%d") != "02-29" & daily$q_m3s > 0, ]
h$doy <- doy_of(h$date)
h$year <- as.integer(format(h$date, "%Y"))
h <- h[h$doy >= season[1] & h$doy <= season[2], ]

# the same baseline years as the window means: a year counts at a station when
# one of its windows has the share of days wet_window_stats() requires
hb <- h[h$status == "approved" & paste(h$station_number, h$year) %in% paste(qb$station_number, qb$year), ]
band <- do.call(rbind, lapply(split(hb, list(hb$station_number, hb$doy), drop = TRUE), function(d) {
  q <- stats::quantile(d$q_m3s, c(0.25, 0.5, 0.75), names = FALSE)
  data.frame(station_number = d$station_number[1], doy = d$doy[1], n = nrow(d),
             lo = q[1], med = q[2], hi = q[3])
}))
band <- band[band$n >= min_baseline, ]
lines <- h[h$year %in% recent, ]
# every day of the season, so a gap in the record breaks the line rather than bridging it
grid <- expand.grid(station_number = stations, year = recent, doy = season[1]:season[2],
                    stringsAsFactors = FALSE)
lines <- merge(grid, lines[c("station_number", "year", "doy", "q_m3s")], all.x = TRUE)
lines <- lines[order(lines$station_number, lines$year, lines$doy), ]
lines$year <- factor(lines$year, names(pal_year$values))

# the windows as strips below the data, on the same log axis
floor_q <- vapply(stations, function(st) {
  min(band$lo[band$station_number == st], lines$q_m3s[lines$station_number == st], na.rm = TRUE)
}, 0)
strip <- do.call(rbind, lapply(stations, function(st) {
  do.call(rbind, lapply(seq_len(nrow(windows)), function(k) {
    data.frame(station_number = st, window = windows$window[k],
               x = doy_of(as.Date(paste0("2001-", windows$start[k]))),
               xend = doy_of(as.Date(paste0("2001-", windows$end[k]))),
               y = floor_q[[st]] / 1.6^k)
  }))
}))
month_start <- doy_of(as.Date(sprintf("2001-%02d-01", 5:10)))
cap_hydro <- sprintf(paste(
  "Daily flow, May to October, %s (lines, broken where a day has no record), over the %s median",
  "day and middle half (drawn only on days with at least %d years). The Chinook windows run",
  "beneath. Log scale."
), recent_text, base_text, min_baseline)
```

The same five years, day by day. The grey band is the middle half of
`r base_text` flows on each day, and its line is the median, not the mean.

```{r hydrograph, fig.height = 6.5, fig.cap = cap_hydro}
ggplot() +
  geom_ribbon(data = by_station(band), aes(doy, ymin = lo, ymax = hi, fill = "band")) +
  geom_line(data = by_station(band), aes(doy, med, colour = "median"), linewidth = 0.5) +
  geom_line(data = by_station(lines), aes(doy, q_m3s, colour = year, group = year), linewidth = 0.45) +
  geom_segment(data = by_station(strip), aes(x = x, xend = xend, y = y, yend = y),
               colour = pal_hydro$values[["window"]], linewidth = 1.8) +
  geom_text(data = by_station(strip), aes(x = season[1] - 3, y = y, label = window), hjust = 1,
            size = 2.7) +
  facet_wrap(~ station_number, ncol = 1, scales = "free_y", labeller = as_labeller(station_label)) +
  # no ticks among the strips: they are rows, not flows
  scale_y_log10(breaks = function(l) {
    b <- 10^(floor(log10(l[1])):ceiling(log10(l[2])))
    b[b >= l[1] * 1.6^(nrow(windows) + 0.5)]
  }, labels = function(x) format(x, drop0trailing = TRUE, scientific = FALSE)) +
  scale_x_continuous(breaks = month_start, labels = month.abb[5:10],
                     limits = c(season[1] - 45, season[2]), expand = c(0, 0)) +
  scale_fill_manual(values = pal_hydro$values["band"], labels = pal_hydro$labels["band"], name = NULL) +
  scale_colour_manual(values = c(pal_hydro$values["median"], pal_year$values),
                      labels = c(pal_hydro$labels["median"], pal_year$labels), name = NULL) +
  labs(x = NULL, y = "Discharge (m³/s)") +
  guides(fill = guide_legend(order = 1), colour = guide_legend(order = 2, nrow = 1)) +
  theme(panel.grid.major.x = element_blank(), legend.box = "vertical",
        legend.spacing.y = unit(0, "pt"), legend.margin = margin(0, 0, 0, 0))
```

## Every year since `r min(baseline)`

```{r heat-data}
heat <- merge(stats[stats$variable == "q_mean" & stats$year >= min(baseline),
                    c("station_number", "period", "year")],
              anom[anom$variable == "q_mean", c("station_number", "period", "year", "anomaly")],
              all.x = TRUE)
heat$class <- factor(as.character(cut(heat$anomaly, c(-Inf, -75, -50, -25, 0, 25, 50, 100, Inf),
                                      labels = paste0("d", 1:8))), names(pal_dep$values))
gap <- setdiff(min(baseline):last_year, heat$year[heat$station_number == "08EE003"])
cap_heat <- sprintf(paste(
  "Window mean flow above or below its %s mean, in %%. Blank cells have under %s of their days,",
  "or none."
), base_text, pct(100 * formals(wet_window_stats)$min_frac))
```

The same departures for every year since `r min(baseline)`, to put the last five
in context. `r station_label[["08EE003"]]` has no window with `r need_text` of
its days in `r years_text(gap)`.

```{r heat, fig.height = 3.6, fig.cap = cap_heat}
ggplot(by_window(by_station(heat), rev = TRUE), aes(year, period)) +
  geom_tile(aes(fill = class), colour = "white", linewidth = 0.2) +
  facet_wrap(~ station_number, ncol = 1, labeller = as_labeller(station_label)) +
  scale_fill_manual(values = pal_dep$values, labels = pal_dep$labels, name = NULL, drop = FALSE,
                    na.translate = FALSE) +
  scale_x_continuous(expand = c(0, 0)) +
  labs(x = NULL, y = NULL) +
  guides(fill = guide_legend(nrow = 2)) +
  theme(panel.grid = element_blank())
```

## Trends

```{r trend-data}
tr <- trend[trend$variable %in% c("q_mean", "q_min7"), ]
tr$variable <- factor(tr$variable, c("q_mean", "q_min7"), c("Window mean", "Lowest 7-day mean"))
tr$trend_start <- factor(tr$trend_start, names(pal_trend$values))
tr$sig <- factor(ifelse(tr$mk_pvalue < alpha, "sig", "ns"), c("sig", "ns"))
n_tested <- nrow(tr)
n_sig <- sum(tr$sig == "sig")
n_neg <- sum(tr$sig == "sig" & tr$slope < 0)
sign_text <- if (n_sig == 0) {
  ""
} else if (n_neg == n_sig) {
  ", all of them negative"
} else {
  sprintf(", %d of them negative", n_neg)
}
# days each pair of windows shares, for the sentence on independence
shared <- length(intersect(days_of(which(windows$window == "Fry migration")),
                           days_of(which(windows$window == "Spawning"))))
stopifnot(shared > 0)   # the sentence calls the tests not independent because of it
# where 08EE003's "from 2000" trends really start: the first year of each fitted series
a3 <- anom[anom$station_number == "08EE003" & anom$year >= trend_start[2] & !is.na(anom$anomaly) &
             anom$variable %in% c("q_mean", "q_min7"), ]
start_003 <- sort(unique(tapply(a3$year, paste(a3$period, a3$variable), min)))
stopifnot(length(start_003) > 0, all(start_003 > trend_start[2]))
cap_trend <- sprintf(paste(
  "Theil-Sen slope of the departure, in percentage points of normal per year, from %s.",
  "Filled where the Mann-Kendall p is below %s."
), paste(trend_start, collapse = " and from "), alpha)
```

The slope of each window's departure, `r trend_text`, for the window mean and
its lowest seven-day mean. Of `r n_tested` slopes, `r n_sig` are significant at
p < `r alpha``r sign_text`. Chance alone would give about
`r round(alpha * n_tested)`, but the tests are not independent: fry migration
shares `r shared` of its days with spawning, and one station drains into the
other. At `r station_label[["08EE003"]]`, "from `r trend_start[2]`" starts in
`r and_text(start_003)`.

```{r trend, fig.height = 4, fig.cap = cap_trend}
ggplot(by_window(by_station(tr), rev = TRUE), aes(slope, period, colour = trend_start, shape = sig)) +
  geom_vline(xintercept = 0, colour = pal_hydro$values[["median"]], linewidth = 0.3) +
  geom_point(position = position_dodge(width = 0.6), size = 2) +
  facet_grid(station_number ~ variable,
             labeller = labeller(station_number = as_labeller(sub(" \\(", "\n(", station_label)))) +
  scale_colour_manual(values = pal_trend$values, labels = pal_trend$labels, name = NULL) +
  scale_shape_manual(values = c(sig = 16, ns = 1),
                     labels = c(sig = paste("p <", alpha), ns = paste("p ≥", alpha)), name = NULL) +
  labs(x = "Slope (percentage points of normal per year)", y = NULL) +
  theme(strip.text.y = element_text(angle = 0, hjust = 0))
```

## What it does not show

```{r limits}
hydat_end <- max(daily$date[daily$source == "hydat"])
n_prov <- length(unique(kf$year[kf$status == "provisional"]))
stopifnot(n_prov > 0)   # the bullet says the latest years are provisional
src <- unique(ch$source)
# the sentence names one source for all three windows, and the one window whose
# citation the table carries over from another watershed's row
stopifnot(identical(src, "gottesfeld_rabnett2007SkeenaFish"))
borrowed <- windows$window[grepl("other area", ch$note)]
stopifnot(identical(borrowed, "Fry migration"))
```

- **The latest years are provisional.** Approved HYDAT flows end in
  `r format(hydat_end, "%B %Y")`, so `r n_prov` of the five years shown rest on
  provisional flows that may still be revised.
- **Some bars have days missing.** A window counts when `r need_text` of its
  days have a recorded flow. `r nrow(part)` of the `r nrow(kf)` bars rest on
  `r pct(part_share[1])`–`r pct(part_share[2])` of their days, where the
  record has gaps: `r part_text`.
- **Some baselines have years missing.** `r miss_text`: no window in those
  years has `r need_text` of its days.
- **The windows come from the literature, not from observation.** NGE's
  life-history table cites Gottesfeld and Rabnett (2007) for all three, though
  its `r tolower(borrowed)` citation is carried over from a neighbouring
  watershed's row.

**Cached inputs.** HYDAT release `r prov$hydat`, with provisional and
real-time flows retrieved `r day_text(prov$retrieved)`; life-history
table at commit `r substr(prov$knowledge_sha, 1, 7)`. Regenerated by
`data-raw/station_vignette_data.R`. Map layers from the Freshwater Atlas
(fwapg) and BC Geographic Names, by `data-raw/station_vignette_map.R`.

```{r wordcount, include = FALSE}
f <- knitr::current_input()
if (!is.null(f)) {
  x <- readLines(f, warn = FALSE)
  y <- utils::tail(x, -utils::head(which(x == "---"), 2)[2])   # drop the YAML header
  fence <- cumsum(grepl("^```", y)) %% 2 == 1
  y <- y[!fence & !grepl("^```", y)]
  y <- gsub("`r [^`]*`", "", y)          # inline code is not body prose
  y <- y[!grepl("^\\s*$", y)]
  n <- sum(lengths(strsplit(trimws(y), "\\s+")))
  message("body prose: ", n, " words")
  cap <- 600   # 511 at #29; the Chinook revision aims well under
  if (n > cap) stop("body prose is ", n, " words; the cap is ", cap)
}
```
